---
title: "rcicrely tutorial: assessing reverse-correlation CI reliability"
author: "Manuel Oliveira"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{rcicrely tutorial: assessing reverse-correlation CI reliability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi      = 96
)
```

## 1. What `rcicrely` is — and isn't

In reverse correlation (RC), participants' trial-level choices are
projected onto a pool of noise patterns to reconstruct the mental
representation they used to make those choices — a **classification
image** (CI). You end up with one CI per participant, plus a group-level
CI obtained by averaging.

`infoVal` (Brinkman et al., 2019) tells you whether a CI contains more
signal than chance. It does **not** tell you whether that signal is
stable across producers, replicable on a different subset of the
producers, or distinguishable from another condition's CI.

`rcicrely` fills that gap with a reliability-assessment toolkit that
operates at the **pixel level**, directly on the CI signal. It gives
you two families of answers:

- **Within-condition** (how consistent is a condition's CI across
  producers?): permuted split-half with Spearman–Brown correction,
  leave-one-out sensitivity, ICC(3,1) and ICC(3,k) via direct
  mean-squares — [`rel_split_half()`][rcicrely::rel_split_half],
  [`rel_loo()`][rcicrely::rel_loo], [`rel_icc()`][rcicrely::rel_icc],
  orchestrated by [`run_within()`][rcicrely::run_within].
- **Between-condition** (is condition A's CI distinguishable from
  condition B's?): vectorised Welch pixel t, cluster-based permutation
  with max-statistic family-wise error control, and bootstrap
  representational dissimilarity — [`rel_cluster_test()`][rcicrely::rel_cluster_test],
  [`rel_dissimilarity()`][rcicrely::rel_dissimilarity], orchestrated by
  [`run_between()`][rcicrely::run_between].

Critically, `rcicrely` does not require a second-phase trait-rating
study: the reliability metrics read from the pixel-level signal
produced by the original producers. This sidesteps the two-phase
design that Cone et al. (2020) flagged as problematic.

## 2. Installation

`rcicrely` is distributed on GitHub. The canonical `rcicr` package
(required only for the 2IFC path; Brief-RC is native here) is also
GitHub-only in its current form:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("rdotsch/rcicr")     # 2IFC path
remotes::install_github("olivethree/rcicrely")
```

If you only work with Brief-RC you can skip `rcicr` — the Brief-RC CI
implementation in `rcicrely` is fully native (Schmitz, Rougier &
Yzerbyt, 2024).

## 3. Two input modes

```{r load}
library(rcicrely)
```

Every reliability metric takes a **signal matrix**: pixels ×
participants, base-subtracted. You can get there in two ways.

### 3.1 Mode 1: pre-computed CI images

If you already have individual CIs on disk as PNG/JPEG:

```{r mode1, eval = FALSE}
signal <- load_signal_matrix(
  dir             = "data/cis_condition_A/",
  base_image_path = "data/base.jpg"
)
# or split the steps:
cis    <- read_cis("data/cis_condition_A/")
signal <- extract_signal(cis, base_image_path = "data/base.jpg")
```

### 3.2 Mode 2: from raw response data

If you have trial-level responses and want the package to compute
CIs internally, dispatch by paradigm.

**2IFC** — wraps `rcicr::batchGenerateCI2IFC()`:

```{r mode2-2ifc, eval = FALSE}
res <- ci_from_responses_2ifc(
  responses       = my_responses,
  rdata_path      = "data/rcicr_stimuli.Rdata",
  baseimage       = "base",
  participant_col = "participant_id"
)
signal <- res$signal_matrix
```

**Brief-RC 12** — native, no `rcicr` dependency:

```{r mode2-briefrc, eval = FALSE}
res <- ci_from_responses_briefrc(
  responses       = my_responses,
  rdata_path      = "data/rcicr_stimuli.Rdata",  # noise pool
  base_image_path = "data/base.jpg",
  method          = "briefrc12"
)
signal <- res$signal_matrix
```

### 3.3 Fake data for this vignette

To keep the vignette self-contained, we simulate a small "condition A
vs condition B" dataset where the two conditions have different latent
signals in different face regions. This makes the reliability metrics
produce non-trivial outputs we can inspect.

```{r fake-data}
set.seed(1)
make_condition <- function(region_top, n_pix = 32L * 32L, n_p = 30L,
                           snr = 0.3) {
  noise <- matrix(rnorm(n_pix * n_p, sd = 0.05), n_pix, n_p)
  mask <- if (region_top) {
    c(rep(1, n_pix / 2L), rep(0, n_pix / 2L))
  } else {
    c(rep(0, n_pix / 2L), rep(1, n_pix / 2L))
  }
  out <- noise + snr * outer(mask, runif(n_p, 0.7, 1.3))
  attr(out, "img_dims") <- c(32L, 32L)
  out
}

sig_A <- make_condition(region_top = TRUE,  n_p = 30L)
sig_B <- make_condition(region_top = FALSE, n_p = 30L)
```

## 4. Within-condition reliability

Run all three within-condition metrics at once via
[`run_within()`][rcicrely::run_within]:

```{r run-within}
within_A <- run_within(
  sig_A,
  n_permutations = 500L,   # 2000+ recommended for production
  seed           = 1L,
  progress       = FALSE
)
print(within_A)
```

### 4.1 Permuted split-half (Spearman–Brown)

For each of `n_permutations` iterations: randomly split the producers
into two halves, average each half to get two group vectors, correlate
them. Spearman–Brown correction
$r_{SB} = 2 r_{hh} / (1 + r_{hh})$ projects to the full sample.
For **odd N**, one producer is randomly dropped per permutation
(re-drawn each iteration).

```{r plot-split-half, fig.cap = "Permutation distribution of split-half r."}
plot(within_A$results$split_half)
```

### 4.2 Leave-one-out sensitivity

For each producer, correlate the full-sample group CI with the group CI
computed without them. Outliers are producers whose LOO correlation
sits more than `flag_threshold_sd` (default 2.5) standard deviations
below the mean — removing them noticeably changes the group CI.

```{r plot-loo, fig.height = 5, fig.cap = "Per-producer LOO correlation."}
plot(within_A$results$loo)
```

### 4.3 ICC(3,1) and ICC(3,k)

Two-way mixed model: pixels are **fixed** (the `img_size × img_size`
grid), participants are **random** (drawn from a producer
population). ICC(3,1) is the reliability of a single producer's CI as
an estimate of the group pattern; ICC(3,k) is the reliability of the
group-mean CI across `k` producers — usually the headline number for
a paper.

```{r plot-icc, fig.cap = "ICC variants reported."}
plot(within_A$results$icc)
```

See [`rel_icc()`][rcicrely::rel_icc]'s help page for why we report
ICC(3,\*) rather than ICC(2,\*) by default — pixels aren't a random
sample from a population of pixels, so ICC(2,\*) is mis-specified even
though it numerically approximates ICC(3,\*) at high pixel counts.
ICC(2,\*) is available via `variants = c("2_1", "2_k")` for reviewers
who expect it.

## 5. Between-condition inference

Run both between-condition metrics at once via
[`run_between()`][rcicrely::run_between]:

```{r run-between}
between_AB <- run_between(
  sig_A, sig_B,
  img_dims        = c(32L, 32L),
  n_permutations  = 500L,
  n_boot          = 500L,
  cluster_threshold = 2.0,
  alpha           = 0.05,
  seed            = 1L,
  progress        = FALSE
)
print(between_AB)
```

### 5.1 Cluster-based permutation test

Pixels where `|Welch t| > cluster_threshold` form candidate clusters
via 4-connectivity (the more conservative choice vs. 8-connectivity).
Each cluster's "mass" is the sum of t-values inside it (captures both
extent and magnitude). A null distribution of max-mass is built by
**stratified** label permutation — shuffling condition labels across
producers while preserving `(N_A, N_B)`. An observed cluster's p-value
is the fraction of null max-masses (matching sign) that equal or
exceed it. Maximum-statistic approach gives FWER control.

```{r plot-cluster, fig.cap = "Per-pixel Welch t-map; significant clusters outlined."}
plot(between_AB$results$cluster_test)
```

### 5.2 Representational dissimilarity

Bootstrap the Pearson correlation and Euclidean distance between the
two conditions' group-level signals. Percentile CIs via base R
`quantile()` — no `boot` dependency.

```{r plot-dissim, fig.cap = "Bootstrap distributions for Pearson r and Euclidean distance."}
plot(between_AB$results$dissimilarity)
```

## 6. Brief-RC workflow end-to-end

The end-to-end Brief-RC path, minus the slow stimulus-generation step
(which you'd run once per project using
`rcicr::generateStimuli2IFC()` to build the noise pool).

```{r briefrc-demo, eval = FALSE}
# 1. Generate a shared noise pool (once).
rcicr::generateStimuli2IFC(
  base_face_files     = list(base = "data/base.jpg"),
  n_trials            = 600L,        # becomes the Brief-RC pool size
  img_size            = 256L,
  stimulus_path       = "data/stim",
  label               = "mystudy",
  ncores              = 1L,          # CLAUDE.md §8.4 — macOS-safe
  return_as_dataframe = TRUE,
  seed                = 1L
)

# 2. Collect trial-level responses in the per-trial format
#    (one row per trial, columns: participant_id, trial, stimulus,
#     response in {-1, +1}). Schmitz et al. 2024, §3.1.2.

# 3. Compute individual CIs, extract signal.
res <- ci_from_responses_briefrc(
  responses       = my_brief_rc_responses,
  rdata_path      = list.files("data/stim", pattern = "\\.rdata$",
                               ignore.case = TRUE, full.names = TRUE)[1],
  base_image_path = "data/base.jpg",
  method          = "briefrc12"
)

# 4. Reliability assessment — same functions as any other paradigm.
within  <- run_within(res$signal_matrix,   seed = 1L)
between <- run_between(res_A$signal_matrix, res_B$signal_matrix,
                       seed = 1L)
```

**Brief-RC infoVal** is deferred to a later release — computing it
correctly requires a reference distribution matched to each
participant's trial count, which is non-trivial. The `rel_*`
reliability metrics don't depend on infoVal.

## 7. Interpreting results and common pitfalls

**Split-half r** — think of $r_{hh}$ around 0.3 as weak, 0.5 as
moderate, 0.7+ as strong for producer-level CI consistency. Because
Spearman–Brown projects from half-samples to the full sample, $r_{SB}$
is the number to report when asked "how reliable is the group CI I'm
publishing?"

**LOO flagging** — a flagged producer means their CI sits far from
the group pattern. That doesn't imply bad data; it could be a
genuinely atypical mental representation. Cross-check against
[`rcicrdiagnostics`][rcicrdiagnostics-link] first to rule out
response-coding errors.

**Cluster test** — a significant cluster tells you *where* two
conditions differ, not *how different they are on average*. Pair with
`rel_dissimilarity()` for a scalar magnitude.

**Sample size** — `rcicrely` warns at N < 30 participants per
condition. Cone et al. (2020) recommend N ≥ 60 for stable reliability
assessment. At N = 10 the metrics are computable but the CIs on them
will be wide.

**ICC(3,\*) vs ICC(2,\*)** — numerically close at high pixel counts,
but ICC(3,\*) is the correctly-specified model (pixels fixed,
producers random). Don't swap to ICC(2,\*) because a reviewer is used
to seeing it — use `variants = c("3_1", "3_k", "2_1", "2_k")` to
report both.

## 8. References

- Brinkman, L., Dotsch, R., Zondergeld, J., Aarts, H. (2019).
  Visualising mental representations: a primer on noise-based reverse
  correlation in social psychology. *European Review of Social
  Psychology*, 28(1), 333–361.
- Cone, J., Flaharty, K., & Ferguson, M. J. (2020). Believability of
  evidence matters for correcting social impressions. *PNAS*.
- Dotsch, R. (2023). *rcicr*: Reverse-Correlation Image-Classification
  Toolbox. R package v1.0.1.
- Maris, E., & Oostenveld, R. (2007). Nonparametric statistical
  testing of EEG- and MEG-data. *Journal of Neuroscience Methods*.
- McGraw, K. O., & Wong, S. P. (1996). Forming inferences about some
  intraclass correlation coefficients. *Psychological Methods*.
- Nichols, T. E., & Holmes, A. P. (2002). Nonparametric permutation
  tests for functional neuroimaging: a primer with examples.
  *Human Brain Mapping*.
- Schmitz, M., Rougier, M., & Yzerbyt, V. (2024). Introducing the
  brief reverse correlation: an improved tool to assess visual
  representations. *Behavior Research Methods*.
- Shrout, P. E., & Fleiss, J. L. (1979). Intraclass correlations:
  uses in assessing rater reliability. *Psychological Bulletin*.

[rcicrdiagnostics-link]: https://github.com/olivethree/rcicrdiagnostics
