---
title: "rcicrely user guide"
author: "Manuel Oliveira"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{rcicrely user guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi       = 96
)
options(rcicrely.silence_scaling_warning = TRUE)  # quiet vignette
```

This guide is written for two readers in parallel: the researcher who
wants to know "which function do I call for the question I have?",
and the methodologically-curious user who also wants to know "and why
should I trust what it returns?". Each function gets a short
plain-language description, the actual signature, a worked example,
how to read the output, and the specific footguns to avoid.

## 1. What `rcicrely` is - and isn't

In reverse correlation (RC), participants' trial-level choices project
onto a pool of noise patterns to reconstruct the mental representation
they used. The output is a per-producer **classification image** (CI)
plus a group-level CI obtained by averaging.

`infoVal` (Brinkman et al., 2019) tells you whether a CI contains more
signal than chance. It does **not** tell you whether that signal is
*stable across producers*, *replicable on a different subset of
producers*, or *distinguishable from another condition's CI*.

`rcicrely` fills that gap with a pixel-level reliability toolkit:

- **Within-condition** (how consistent is a condition's CI across
  producers?): permuted split-half with Spearman-Brown,
  leave-one-out sensitivity, ICC(3,1) and ICC(3,k).
- **Between-condition** (is condition A's CI distinguishable from
  condition B's?): vectorised Welch t per pixel, cluster-based
  permutation with max-statistic FWER, bootstrap representational
  dissimilarity.

Critically, none of this requires a phase-2 trait-rating study. The
metrics read directly from the pixel-level signal produced by the
original producers, sidestepping the two-phase design Cone,
Brown-Iannuzzi, Lei, and Dotsch (2021) showed inflates Type I error.

### `infoVal` and `rcicrely` together

Since v0.2 `rcicrely` ships its own `infoval()` (see §6.5) that
handles **both 2IFC and Brief-RC** with a single function. The
reference distribution is keyed on each producer's actual trial
count rather than the full pool size — the calibration matters most
for Brief-RC, where producers see only a subset of the pool, but the
function works equivalently for 2IFC. The two checks remain
complementary: `infoval()` tests presence of signal; `rel_*()` tests
reliability and discriminability of signal.

`rcicr::computeInfoVal2IFC()` is still useful as an external cross-
check for 2IFC data. It pulls the raw `$ci` element from the rcicr
CI-list internally (`norm(matrix(target_ci[["ci"]]), "f")`), so the
scaling settings used at CI generation do not affect its output.

The scaling caveat applies when you compute `infoVal` outside that
path:

- **rcicrely's `infoval()`** expects raw masks. Pass
  `res$signal_matrix` from `ci_from_responses_*()` (Mode 2). Do not
  pass `res$rendered_ci` or anything PNG-derived.
- **Hand-rolled infoVal**: same rule. Use the raw mask, not the
  rendered PNG.

## 2. Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("rdotsch/rcicr")      # 2IFC path only
remotes::install_github("olivethree/rcicrely")
```

If you only use Brief-RC you can skip `rcicr` - the Brief-RC CI
implementation is fully native (Schmitz, Rougier & Yzerbyt, 2024).

```{r load}
library(rcicrely)
```

## 3. Mental model: the signal matrix

Every reliability metric in this package takes a single object: a
**signal matrix**, pixels by participants. Get this right and the
rest of the package is a thin shell over it.

### 3.1 Raw mask vs. rendered CI

A CI exists in two forms.

The **raw mask** is the participant's noise contribution before any
display transformation. It is what the math is about. For Brief-RC,
it is `(noise_pool[, chosen_ids] %*% signed_responses) / n_unique`
(Schmitz's `genMask()`). For 2IFC, it is rcicr's `$ci` element. Values
are typically in a narrow band around zero (~ +/- 0.05).

The **rendered CI** is what gets saved to PNG. It is
`base + scaling(mask)`, where `scaling()` may stretch the mask to the
base image's range ("matched"), apply a constant gain, or do nothing.
Scaling is a *display* transformation, not a statistical one.

Subtracting the base from a rendered CI does **not** recover the raw
mask:

```
PNG_pixels - base = scaling(mask)   (NOT  mask)
```

| Metric                                       | Behavior under scaling                                       |
|---------------------------------------------:|:-------------------------------------------------------------|
| `rel_split_half`, `rel_loo`                  | Survives uniform scaling. Breaks under per-CI "matched".     |
| `rel_dissimilarity` (Pearson r half)         | Same.                                                        |
| `rel_dissimilarity` (Euclidean half)         | **Distorted** by any scaling.                                |
| `rel_icc`                                    | **Distorted** by any scaling.                                |
| `pixel_t_test`, `rel_cluster_test`           | **Distorted** by per-CI scaling.                             |
| `rcicr::computeInfoVal2IFC()` (canonical)    | Uses raw `$ci` internally, unaffected.                    |
| Hand-rolled `infoVal` (Brief-RC, custom)     | **Distorted** by any scaling. Use the raw mask.              |

### 3.2 Two paths to the signal matrix

```
Mode 1: PNGs on disk           Mode 2: raw responses
          |                              |
   read_cis()                    ci_from_responses_2ifc() / _briefrc()
          |                              |
   extract_signal()                      |
          |                              |
   scaling(mask)                       mask                    <-- raw
          |                              |
          +---------> rel_*() <----------+
```

**Mode 2 returns the raw mask**; Mode 1 cannot, because PNGs on disk
have always been rendered. Prefer Mode 2 when raw responses are
available. If you must use Mode 1, the package emits a one-time
warning per session reminding you of the caveat. To silence:

```{r silence-mode1, eval = FALSE}
# acknowledge per-call:
signal <- load_signal_matrix("data/cis/", "data/base.jpg",
                             acknowledge_scaling = TRUE)

# or silence for the whole session:
options(rcicrely.silence_scaling_warning = TRUE)
```

If you generated the PNGs yourself with
`rcicr::generateCI2IFC(scaling = "none")`, the rendered output is
effectively raw, and the warning is genuinely a false positive.

### 3.3 A concrete demonstration

Here is what scaling does to the same underlying mask.

```{r scaling-demo, fig.height = 3.5, fig.width = 7}
set.seed(1)
n_pix <- 32L * 32L
n_p   <- 20L
mask  <- matrix(rnorm(n_pix * n_p, sd = 0.02), n_pix, n_p)

# Render: stretch each column to [-0.5, 0.5] (Schmitz "matched")
rendered <- apply(mask, 2L, function(m) {
  rng <- diff(range(m)); if (rng == 0) m else m * (1 / rng)
})

c(
  raw_max_range      = round(diff(range(mask)),     3L),
  rendered_max_range = round(diff(range(rendered)), 3L)
)

# Pearson r between same-column raw and rendered: ~ 1
# (a perfect linear stretch preserves Pearson)
mean(vapply(seq_len(n_p),
            function(j) cor(mask[, j], rendered[, j]),
            numeric(1L)))

# But the variance ratios that drive ICC are very different
var(rowMeans(mask))      # variance of the group mean over pixels
var(rowMeans(rendered))  # same metric, on the rendered version
```

Pearson r per column is exactly 1 (linear scaling is correlation-
preserving), but the row-mean variance the ICC formula depends on is
~25x larger on the rendered version. ICC(3,*) computed on rendered
data is therefore not the same ICC computed on the raw mask. The
same logic applies to Euclidean distance and to `infoVal`.

### 3.4 A recurring real-world example

The synthetic demos in the function-reference sections below are
deliberately small so the vignette knits in seconds. To see what
each function looks like on a published RC dataset, the sections
that follow each include a short **Real example** callout drawn
from one source:

> Oliveira, M., Garcia-Marques, T., Dotsch, R., & Garcia-Marques,
> L. (2019). Dominance and competence face to face: Dissociations
> obtained with a reverse correlation approach. *European Journal
> of Social Psychology*. <https://doi.org/10.1002/ejsp.2569>

In Study 1, 200 participants completed a 2IFC reverse-correlation
task with **300 trials each** on a 256x256 male base face,
across 10 trait conditions in a between-subjects design (20
producers per trait). Trait labels: `Dominant`, `Submissive`,
`Trust`, `Untrust`, `Friendly`, `Unfriendly`, `Intelligent`,
`Unintelligent`, `Competent`, `Incompetent`. The full data,
parameter file, and analysis scripts are openly available on OSF:
<https://doi.org/10.17605/osf.io/hr5pd>.

The "Real example" callouts below show the exact call you would
make against this data, plus the output captured from a complete
replication run. The replication itself (including reconstructing
the noise matrix from the legacy `rcicr` 0.3.0 parameter file) is
in `study1_replication.Rmd` in the project's `temp/` directory; the
vignette chunks below are marked `eval = FALSE` because the data
is not bundled with the package, but every output shown is real.

## 4. Function reference: I/O

### 4.1 `read_cis()`

**What it does.** Reads every PNG/JPEG in a directory, converts each
to grayscale, returns a pixels by participants numeric matrix.

**When to use it.** You have one CI image per producer already saved
on disk and you want to load them into R as a numeric matrix. Pair
with [`extract_signal()`](#section-extract-signal) to get a
base-subtracted signal matrix.

```{r read-cis-demo}
# Make a tiny synthetic dataset on disk to demonstrate against.
demo_dir <- tempfile("ci_demo_"); dir.create(demo_dir)
base_img <- matrix(0.5, 32L, 32L)
png_path <- function(i) file.path(demo_dir, sprintf("p%02d.png", i))
if (requireNamespace("png", quietly = TRUE)) {
  for (i in 1:6) {
    img <- pmin(pmax(base_img + matrix(rnorm(32 * 32, sd = 0.05), 32, 32),
                     0), 1)
    png::writePNG(img, png_path(i))
  }
}
```

```{r read-cis-call, eval = requireNamespace("png", quietly = TRUE)}
cis <- read_cis(demo_dir, acknowledge_scaling = TRUE)
dim(cis)
attr(cis, "img_dims")
```

**Reading the result.** `nrow = n_pixels`, `ncol = n_files`. The
`img_dims` attribute carries `c(nrow, ncol)` of the original image so
plotting helpers know how to reshape.

**Common mistakes.** Forgetting to subtract the base before passing
to a `rel_*` function (use [`extract_signal()`](#section-extract-signal)
or the convenience wrapper [`load_signal_matrix()`](#section-load-signal-matrix)).
Mixing CIs of different sizes - aborts loudly. Trusting reliability
numbers without reading the raw vs. rendered note (chapter 3).

### 4.2 `extract_signal()` {#section-extract-signal}

**What it does.** Subtracts the base image from each column of a CI
matrix, so reliability metrics see only the participant-specific
contribution.

**When to use it.** You have the output of [`read_cis()`](#read_cis)
and you want a base-subtracted signal matrix.

```{r extract-signal-demo, eval = requireNamespace("png", quietly = TRUE)}
base_path <- file.path(demo_dir, "_base.png")
png::writePNG(base_img, base_path)
signal <- extract_signal(cis, base_image_path = base_path,
                         acknowledge_scaling = TRUE)
dim(signal); attr(signal, "img_dims")
```

**Reading the result.** Same shape as `cis`, with the `img_dims`
attribute preserved. Column names propagate.

**Common mistakes.** Pointing at the wrong base (a generated CI PNG
of the right size will pass the dimension check). Treating the result
as the raw mask; it is `scaling(mask)` if PNGs were rendered with any
scaling. See chapter 3.

### 4.3 `load_signal_matrix()` {#section-load-signal-matrix}

**What it does.** Convenience wrapper that calls [`read_cis()`](#read_cis)
and [`extract_signal()`](#section-extract-signal) in one go.

**When to use it.** Most of the time, when you don't need to do
anything between reading and subtracting (e.g. masking pixels,
swapping the base).

```{r load-signal-demo, eval = requireNamespace("png", quietly = TRUE)}
signal <- load_signal_matrix(
  dir                 = demo_dir,
  base_image_path     = base_path,
  pattern             = "^p[0-9]+\\.png$",  # exclude _base.png
  acknowledge_scaling = TRUE
)
dim(signal)
```

### 4.4 `read_noise_matrix()`

**What it does.** Loads the pixels by trials noise pool used at
stimulus generation. Three formats: rcicr `.Rdata` (reconstructs from
`stimuli_params` + `p`), saved `.rds`, or whitespace-delimited text.

**When to use it.** You are computing Brief-RC CIs and want to load
the noise pool once, then reuse it across multiple
[`ci_from_responses_briefrc()`](#section-cifr-briefrc) calls.

```{r read-noise-demo, eval = FALSE}
# from the rcicr rdata (slow - reconstructs each trial via
# rcicr::generateNoiseImage):
noise <- read_noise_matrix("data/rcicr_stimuli.Rdata")

# from a cached rds you saved from the rcicr return value (fast):
noise <- read_noise_matrix("data/noise_matrix.rds")
```

**Common mistakes.** Assuming the rcicr `.Rdata` contains a `stimuli`
object (it does not - the pixels-by-trials matrix is the *return
value* of `generateStimuli2IFC(return_as_dataframe = TRUE)`, never
saved to the file). Cache to `.rds` once and load from there.
Multiple stale `.rdata` files in one directory: rcicr timestamps
filenames; pick the most recent by mtime if you have several.

### 4.5 `ci_from_responses_2ifc()` {#section-cifr-2ifc}

**What it does.** Wraps `rcicr::batchGenerateCI2IFC()` with all known
gotchas pre-handled, returns a uniform result list.

**When to use it.** You have 2IFC trial-level responses and want
per-producer CIs in one call. Requires `rcicr` installed.

```{r cifr-2ifc-demo, eval = FALSE}
res <- ci_from_responses_2ifc(
  responses       = my_responses,
  rdata_path      = "data/rcicr_stimuli.Rdata",
  baseimage       = "base",
  participant_col = "participant_id",
  keep_rendered   = FALSE   # set TRUE if you want $rendered_ci for plotting
)
signal <- res$signal_matrix
```

**Reading the result.** `$signal_matrix` is the **raw mask** (rcicr's
`$ci` per producer) - feed this to `rel_*`. `$participants` and
`$img_dims` are convenience metadata. `$rcicr_result` is the raw
rcicr return value if you want anything else. `$rendered_ci`
(present only if `keep_rendered = TRUE`) is `base + scaling(mask)`
per producer - **visualization only**, do not feed to `rel_*`.

**Common mistakes.** Inventing a `base_image_path` argument - the base
is read from the rdata via `baseimage` *label*, not a path. Passing
`$rendered_ci` to `rel_*`. Trusting `scaling = "autoscale"` to be
"raw" - the rendered output is scaled even when the signal_matrix
isn't.

### 4.6 `ci_from_responses_briefrc()` {#section-cifr-briefrc}

**What it does.** Native Brief-RC 12 implementation of Schmitz's
`genMask()` formula. No `rcicr::*_brief` calls (those don't exist in
canonical rcicr v1.0.1).

**When to use it.** You have Brief-RC 12 trial-level responses and
the noise pool used at stimulus generation.

```{r cifr-briefrc-demo, eval = FALSE}
res <- ci_from_responses_briefrc(
  responses       = my_responses,    # one row per trial
  rdata_path      = "data/rcicr_stimuli.Rdata",
  base_image_path = "data/base.jpg",
  method          = "briefrc12",
  scaling         = "none"           # raw mask only - no $rendered_ci
)
signal <- res$signal_matrix
```

To also get the rendered CI for plotting:

```{r cifr-briefrc-rendered, eval = FALSE}
res <- ci_from_responses_briefrc(
  responses       = my_responses,
  noise_matrix    = my_noise,
  base_image_path = "data/base.jpg",
  scaling         = "matched"        # Schmitz default for visualisation
)
# res$signal_matrix is the raw mask, feed to rel_*
# res$rendered_ci is base + matched(mask), for PNG / plotting only
```

**Reading the result.** `$signal_matrix` is **always** the raw mask,
regardless of `scaling`. `$rendered_ci` exists only when
`scaling != "none"`. `$participants`, `$img_dims`, `$scaling` are
metadata.

**Common mistakes.** Passing the "expanded" 12-rows-per-trial format -
Brief-RC 12 is one row per trial (the chosen face), not 12. Asking
for `method = "briefrc20"` (deferred). Using `$rendered_ci` for
downstream stats.

```{r cleanup-demo, include = FALSE}
unlink(demo_dir, recursive = TRUE)
```

## 5. Function reference: within-condition reliability

For the rest of this guide we'll work with a small synthetic dataset
where two conditions have different latent signals.

```{r fake-data}
set.seed(1)
make_condition <- function(region_top, n_pix = 32L * 32L,
                           n_p = 30L, snr = 0.3) {
  noise <- matrix(rnorm(n_pix * n_p, sd = 0.05), n_pix, n_p)
  mask <- if (region_top) {
    c(rep(1, n_pix / 2L), rep(0, n_pix / 2L))
  } else {
    c(rep(0, n_pix / 2L), rep(1, n_pix / 2L))
  }
  out <- noise + snr * outer(mask, runif(n_p, 0.7, 1.3))
  attr(out, "img_dims") <- c(32L, 32L)
  out
}
sig_A <- make_condition(region_top = TRUE,  n_p = 30L)
sig_B <- make_condition(region_top = FALSE, n_p = 30L)
```

### 5.1 `rel_split_half()` {#section-split-half}

**What it does.** Estimates how stable a condition's group CI is
across random halves of producers. Reports the per-permutation
correlation between halves and the Spearman-Brown projection to the
full sample.

**When to use it.** You want to answer "would I get the same group
CI if I'd run a different half of my participants?". `r_SB` is
usually the headline number.

```{r rel-split-half}
r_sh <- rel_split_half(sig_A, n_permutations = 500L, seed = 1L,
                       progress = FALSE)
print(r_sh)
```

```{r plot-split-half, fig.cap = "Permutation distribution of split-half r."}
plot(r_sh)
```

**Reading the result.** `$r_hh` = mean per-split r. `$r_sb` =
Spearman-Brown projected reliability of the full sample (report this
unless asked otherwise). `$ci_95`, `$ci_95_sb` are percentile CIs.
`$distribution` is the full per-permutation r vector.

**Common mistakes.** Reporting `$r_hh` instead of `$r_sb` when the
audience cares about the full-sample CI. Running with
`n_permutations < 200` and trusting the CIs.

**Real example (Oliveira et al., 2019, Study 1).** Per-trait
group-CI reliability for the Trust and Dominant conditions (20
producers each, 300 trials, 256x256 image):

```{r split-half-oliveira-2019, eval = FALSE}
# sm_trust, sm_dominant: pixels x 20 producer signal matrices
# (built from the response data; see study1_replication.Rmd)
rel_split_half(sm_trust,    n_permutations = 500L, seed = 1L)
#> Trust:    r_hh = 0.405, r_sb = 0.577, 95% CI on r_sb = [0.539, 0.616]
rel_split_half(sm_dominant, n_permutations = 500L, seed = 1L)
#> Dominant: r_hh = 0.262, r_sb = 0.415, 95% CI on r_sb = [0.380, 0.450]
```

The Trust CI replicates with a clearly higher group-level
reliability than the Dominant CI, consistent with the inter-rater
agreement ordering reported in Table 1 of Oliveira et al. (2019).

### 5.2 `rel_loo()` - influence screening (not a reliability metric) {#section-rel-loo}

**What it does.** For each producer, correlates the full-sample
group CI with the group CI computed without that producer, and
returns a **z-scored** version of those correlations. Producers
whose z-score sits clearly below zero have a disproportionate
influence on the group pattern and are flagged for inspection.

**When to use it.** To screen for producers whose individual CI is
far from the group pattern - often a sign of response miscoding,
fatigue, or a genuinely atypical mental representation. This is a
diagnostic; it is *not* a reliability statistic (see the note at the
end of this section).

```{r rel-loo}
r_loo <- rel_loo(sig_A, flag_threshold = 2.5, flag_method = "sd")
print(r_loo)
```

The default is mean +/- SD; use `flag_method = "mad"` for a
median +/- MAD rule that doesn't get pulled around by the very
outliers it tries to detect:

```{r rel-loo-mad}
r_loo_mad <- rel_loo(sig_A, flag_threshold = 2.5, flag_method = "mad")
c(sd_threshold = r_loo$threshold, mad_threshold = r_loo_mad$threshold)
```

```{r plot-loo, fig.height = 5, fig.cap = "z-scored LOO influence per producer."}
plot(r_loo)
```

For a tidy ordered table of producers by influence, use
`rel_loo_z()`:

```{r rel-loo-z, eval = FALSE}
head(rel_loo_z(r_loo))
#   participant_id  correlation  z_score  flag
# 1     outlier01        0.9714    -3.12  TRUE
# 2           p07        0.9942    -0.84  FALSE
# ...
```

**Reading the result.** `$z_scores` is the informative quantity -
plot / report it directly. `$correlations` is retained for
transparency (values near 1 by construction because the full-sample
mean and the leave-one-out mean share `(N - 1) / N` of their data).
`$flagged` lists ids below threshold. `$summary_df` is a tidy table
sorted by `z_score`.

**Why you should not read `$correlations` as reliability.** `r_loo`
values are almost always in `[0.95, 0.999]` at N = 30, even on very
noisy CIs, because the two vectors being correlated overlap in
`(N - 1) / N` of their data. A naive reader can misinterpret
"LOO r = 0.98" as "the CIs are 98% reliable"; this is wrong. Report
reliability via `rel_split_half()` or `rel_icc()` (§5.1, §5.3).

**Common mistakes.** Treating `$flagged` as "drop these producers" -
investigate first. A flagged producer may have a genuinely atypical
mental representation rather than bad data. Cross-check with
`rcicrdiagnostics` to rule out response-coding errors.

**Real example (Oliveira et al., 2019, Study 1, Trust condition).**
Median raw `r_loo` is .992 (MAD = .001) — illustrating the
near-1 absolute value structurally. Zero producers cleared the
default MAD-based flag threshold, but the z-scored ordering is
still informative for inspection priority:

```{r loo-oliveira-2019, eval = FALSE}
loo <- rel_loo(sm_trust, flag_method = "mad")
head(rel_loo_z(loo), 3)
#>   participant_id  correlation  z_score  flag
#> 1          18043       0.9905    -1.68  FALSE
#> 2          18051       0.9911    -0.99  FALSE
#> 3          18131       0.9913    -0.79  FALSE
```

The producers at the top of the z-ordering are the ones whose
removal moves the group CI most relative to the rest — even when
none clear the flag threshold, this is the inspection order.

### 5.3 `rel_icc()` {#section-rel-icc}

**What it does.** Two-way mixed ICC via direct mean squares. Default
returns ICC(3,1) and ICC(3,k). Pixels are fixed (the image grid),
participants are random (drawn from a producer population).

**When to use it.** Single-scalar reliability summary. ICC(3,k) is
"how stable is the group-mean CI?", ICC(3,1) is "how informative is
one producer's CI?".

```{r rel-icc}
r_icc <- rel_icc(sig_A, variants = c("3_1", "3_k"))
print(r_icc)
```

To also report ICC(2,*) for a reviewer who expects it:

```{r rel-icc-both}
r_icc_full <- rel_icc(sig_A, variants = c("3_1", "3_k", "2_1", "2_k"))
print(r_icc_full)
```

**Reading the result.** `$icc_3_1`, `$icc_3_k`, `$icc_2_1`,
`$icc_2_k` (the latter two NA unless requested). `$ms_rows`,
`$ms_cols`, `$ms_error` are the underlying ANOVA mean squares for
transparency. `$model` is a description string the print method shows.

**Common mistakes.** Asking for ICC(2,*) because a reviewer expects it.
ICC(2,*) treats pixels as a random sample from a pixel population,
which they aren't. The numbers are usually close to ICC(3,*) at high
pixel counts but the model is mis-specified - report both with
`variants = c("3_1", "3_k", "2_1", "2_k")` if asked.

ICC is **strongly** scale-sensitive. Use the raw mask only.

**Real example (Oliveira et al., 2019, Study 1).** Two-way mixed
ICC on 65,536-pixel raw masks, 20 producers per condition:

```{r icc-oliveira-2019, eval = FALSE}
rel_icc(sm_trust)
#> ICC(3,1) = 0.1008,  ICC(3,k) = 0.6917
rel_icc(sm_dominant)
#> ICC(3,1) = 0.0489,  ICC(3,k) = 0.5068
```

ICC(3,1) — the resolution-comparable single-producer reliability —
is clearly higher for Trust than for Dominant, matching the
reliability ordering reported in the original paper. ICC(3,k) is
high in absolute terms but the resolution asymptote is a known
property of ICC(*,k) at large pixel counts; report ICC(3,1) as the
primary cross-study comparable number.

### 5.4 `run_within()` {#section-run-within}

**What it does.** Calls all three within-condition metrics and bundles
the result for joint printing/plotting.

```{r run-within}
within_A <- run_within(
  sig_A,
  n_permutations = 500L,
  seed           = 1L,
  progress       = FALSE
)
print(within_A)
```

`$results$split_half` and `$results$icc` are the two reliability
result objects. `plot(within_A)` lays them out side by side.
Leave-one-out influence screening is run separately with
`rel_loo()` — it is an influence diagnostic, not a reliability
statistic (see §5.2).

## 6. Function reference: between-condition inference

### 6.1 `pixel_t_test()` {#section-pixel-t}

**What it does.** Vectorised Welch's t per pixel between two
condition signal matrices.

**When to use it.** You want the raw t-map (e.g. for custom plotting,
threshold sensitivity analysis) without cluster-level inference.

```{r pixel-t}
t_vec <- pixel_t_test(sig_A, sig_B)
range(t_vec); length(t_vec)
```

Welch's t (unequal variances) is correct - 2IFC and Brief-RC
conditions can differ in N and variance.

### 6.2 `rel_cluster_test()` {#section-cluster}

**What it does.** Cluster-based permutation test with max-statistic
FWER control. Pixels above the threshold form clusters via
4-connectivity; cluster mass = sum of t-values inside; null built by
**stratified** label permutation; max-mass approach controls
family-wise error.

**When to use it.** "Where in the image do these two conditions
differ?" with proper multiple-comparison control across pixels.

```{r rel-cluster}
clust <- rel_cluster_test(
  sig_A, sig_B,
  img_dims          = c(32L, 32L),
  n_permutations    = 500L,
  cluster_threshold = 2.0,
  alpha             = 0.05,
  seed              = 1L,
  progress          = FALSE
)
print(clust)
```

```{r plot-cluster, fig.cap = "Per-pixel Welch t-map; significant clusters outlined."}
plot(clust)
```

**Reading the result.** `$observed_t` is the raw t-map.
`$clusters` is a data.frame with `cluster_id`, `direction`, `mass`,
`size`, `p_value`, `significant`. `$pos_labels` / `$neg_labels` are
integer matrices for plotting overlays. `$null_distribution` exposes
the per-permutation max masses.

**Common mistakes.** Reading `cluster_threshold = 2.0` as "p < 0.05"
- it is the t-cutoff for forming candidate clusters, not the cluster
significance threshold. Significance is the cluster's mass relative
to the null. Trusting cluster p-values with `n_permutations < 1000`.

**Real example (Oliveira et al., 2019, Study 1).** A central claim
of the original paper was that the Trust and Friendly CIs overlap
strongly while the Dominant and Competent CIs do not. Both threshold
and TFCE versions of the cluster test, run on the per-producer
signal matrices for the relevant pairs, support this in
spatially-localized terms (FWER-controlled). The full output is in
`temp/study1_replication.Rmd`; the runnable form is:

```{r cluster-oliveira-2019, eval = FALSE}
rel_cluster_test(sm_dominant, sm_competent,
                 img_dims = c(256L, 256L),
                 method   = "tfce",
                 n_permutations = 500L,
                 tfce_n_steps   = 50L,
                 seed = 1L)
```

Pair this with `rel_dissimilarity()` (next section) for the
overall-magnitude part of the same claim.

### 6.3 `rel_dissimilarity()` {#section-dissim}

**What it does.** Two scalar dissimilarity metrics between the two
conditions' group CIs - Pearson correlation and Euclidean distance -
each with percentile bootstrap CIs.

**When to use it.** "How different are these two conditions overall?"
with uncertainty. Pair with `rel_cluster_test()` for spatial
information.

```{r rel-dissim}
diss <- rel_dissimilarity(
  sig_A, sig_B,
  n_boot   = 500L,
  ci_level = 0.95,
  seed     = 1L,
  progress = FALSE
)
print(diss)
```

```{r plot-dissim, fig.cap = "Bootstrap distributions for Pearson r and Euclidean distance."}
plot(diss)
```

**Reading the result.** `$correlation`, `$euclidean` are the
observed values; `$ci_cor`, `$ci_dist` the percentile CIs;
`$boot_cor`, `$boot_dist` the full bootstrap distributions.

**Common mistakes.** Reading `$correlation` near 1 as "the conditions
look the same" - identical spatial patterns with different magnitudes
give r ~ 1; the magnitude difference shows up in `$euclidean`.
`$euclidean` scales with sqrt(n_pixels) so it is a within-study
metric, not comparable across resolutions.

**Real example (Oliveira et al., 2019, Study 1).** Euclidean
distances between group-mean CIs on 256x256 raw masks, 500
bootstrap replicates each:

```{r dissim-oliveira-2019, eval = FALSE}
rel_dissimilarity(sm_trust,    sm_friendly, n_boot = 500L, seed = 1L)
#> Euclidean = 16.81, 95% CI [19.26, 24.86], normalised = 0.0657
rel_dissimilarity(sm_dominant, sm_competent, n_boot = 500L, seed = 2L)
#> Euclidean = 21.08, 95% CI [21.53, 30.75], normalised = 0.0823
```

The Dominant-vs-Competent contrast yields a substantially larger
Euclidean distance than Trust-vs-Friendly, the same direction as
the published group-CI Pearson correlations (.69 for Trust x
Friendly, -.22 for Dominant x Competent in Table 2 of the paper).
The two metrics report the same pattern from different angles —
Pearson is the within-shape question, Euclidean is the
how-far-apart question.

### 6.4 `run_between()` {#section-run-between}

**What it does.** Calls both between-condition metrics and bundles
the result.

```{r run-between}
between_AB <- run_between(
  sig_A, sig_B,
  img_dims        = c(32L, 32L),
  n_permutations  = 500L,
  n_boot          = 500L,
  seed            = 1L,
  progress        = FALSE
)
print(between_AB)
```

### 6.5 `infoval()` - producer-level signal vs. chance {#section-infoval}

**What it does.** For each producer, compares the Frobenius norm of
their raw mask to a reference distribution simulated from random
sign-weighted aggregations of the same noise pool, at that
producer's trial count. Returns a per-producer z-score against the
reference's median / MAD, plus the full reference distribution and
metadata. A single function handles both 2IFC and Brief-RC; the
difference lives entirely in what you pass as `noise_matrix`.

**When to use it.** As a complement to the `rel_*()` reliability
checks. `infoval()` answers *does each producer's mask carry more
signal than chance?*; `rel_*()` answers *how reliable / how
distinguishable is the group-level pattern?*. They report different
things and a study should usually carry both.

```{r infoval-example, eval = FALSE}
# Brief-RC route: infoval() takes the raw signal_matrix from
# ci_from_responses_briefrc() and the noise_matrix used to generate
# the trials.
res <- ci_from_responses_briefrc(responses, rdata_path = "rcic.Rdata",
                                 base_image_path = "base.jpg")
trial_counts <- with(responses,
                     setNames(table(participant_id),
                              names(table(participant_id))))
iv <- infoval(res$signal_matrix, noise_matrix,
              trial_counts = trial_counts,
              iter = 10000L, cache_path = "infoval_cache.rds",
              seed = 1L)
print(iv)
```

**Reading the result.** `$infoval` is a per-producer z-score vector;
`z > 1.96` is roughly one-tailed *p* < .025 against the chance
reference. `$norms` is the raw observed Frobenius norm per producer.
`$reference` holds the simulated reference norm vectors keyed by
trial count. `$ref_median` and `$ref_mad` are the centre / spread
used for z computation.

#### Troubleshooting low or negative `infoval` z-scores

If you compute `infoval()` and find that *most* or *all* per-producer
z-scores sit well below 1.96 — sometimes negative — even though spot
checks suggest producers are doing the task seriously, that is a
common pattern, not a sign the data is bad. Five reasons in order
of how often they apply:

1. **Frobenius norm is a global energy statistic.** It sums squared
   pixel deviations across the *entire* image. Real internal
   representations are usually spatially sparse (eyes, mouth, jaw —
   maybe 10-30% of pixels), so the 70-90% of "background" pixels
   contribute noise of similar magnitude to the chance reference and
   dilute the signal-bearing region. A producer with strong,
   visually-obvious signal in the eyes can have a Frobenius norm
   only marginally above the random reference.

2. **The reference is strict because it lives in the same subspace.**
   Both the observed mask and the reference are projections onto the
   same low-dimensional sinusoidal noise basis. The reference
   distribution has plenty of overall energy by construction, so the
   only way to clear z = 1.96 is to align signs with a *specific
   subset* of patterns more than chance.

3. **Per-trial signal is small.** Each 2IFC choice contributes a
   tiny signal increment relative to the per-trial noise amplitude.
   With 300 trials the SNR gain is √300 ≈ 17×, but if per-trial
   signal is ~5% of per-trial noise, post-aggregation effective SNR
   is barely visible to a global energy measure.

4. **Without a face mask, infoVal counts background.**
   `face_mask()` ships an oval mask that approximates the Schmitz
   2024 face region. Applying it (`infoval(..., mask =
   face_mask(c(256, 256)))`) concentrates the norm on signal-bearing
   pixels and typically lifts z-scores noticeably.

5. **Group-level CIs have much higher z than individual CIs.**
   Averaging 20 producers' masks reduces noise by √20 ≈ 4.5×, so
   the group-mean CI's effective trial count is `300 × 20 = 6000`
   for a 20-producer condition. The infoVal of the group-mean CI
   (with `trial_counts = setNames(300L * n_producers, "group")`)
   is usually 5-10× the per-producer median. **This is what
   Brinkman et al. (2019) reported.** Per-producer infoVal is
   noisier and lower; this is structural, not a defect.

##### Diagnostic recipe

```{r infoval-diagnostic, eval = FALSE}
sm     <- res$signal_matrix
tc     <- setNames(rep(300L, ncol(sm)), colnames(sm))

# 1. Compare the observed and reference norm distributions
iv  <- infoval(sm, noise_matrix, tc, iter = 1000L,
               cache_path = "iv.rds", seed = 1L)
ref <- iv$reference[[as.character(tc[1])]]
cat(sprintf("observed median = %.4f, reference median = %.4f, %% above = %+.1f%%\n",
            median(iv$norms), median(ref),
            100 * (median(iv$norms) - median(ref)) / median(ref)))

# 2. Apply the face mask
fm  <- face_mask(c(256L, 256L))
iv_masked <- infoval(sm, noise_matrix, tc, mask = fm,
                     iter = 1000L, cache_path = "iv_masked.rds",
                     seed = 1L)
median(iv_masked$infoval)        # typically higher than iv

# 3. Group-mean CI (the headline number to report)
group   <- matrix(rowMeans(sm), ncol = 1, dimnames = list(NULL, "group"))
tc_grp  <- setNames(300L * ncol(sm), "group")
iv_grp  <- infoval(group, noise_matrix, tc_grp, iter = 1000L,
                   seed = 1L)
iv_grp$infoval                   # usually large (e.g. > 5)

# 4. Sanity check with a simulated random responder
set.seed(0)
random_resp <- sample(c(-1, 1), 300L, replace = TRUE)
random_mask <- (noise_matrix %*% random_resp) / 300
iv_rand <- infoval(matrix(random_mask, ncol = 1,
                          dimnames = list(NULL, "rand")),
                   noise_matrix,
                   setNames(300L, "rand"),
                   iter = 1000L, seed = 1L)
iv_rand$infoval                  # should be ~ 0 within MAD noise
```

**Negative z values** mean the observed mask has *less* total
Frobenius energy than the chance reference. This is not "negative
signal" — it is consistent with a producer whose responses align
with a structured low-rank template, since aligning consistently
with one template can reduce the effective dimensionality of the
mask and shrink its overall norm. A negative z is therefore
informative: a clearly negative z (say, < −2) on a producer who
otherwise shows good behaviour usually points to one of:

- The producer responded coherently but to a *different* feature
  than the trait you intended (idiosyncratic representation).
- The reference distribution is mis-calibrated for your design (see
  the v0.2.1 calibration note in the `?infoval` help page —
  early v0.2.0 versions sampled stim ids with replacement and
  inflated reference norms, biasing per-producer z below zero on
  cooperative producers).
- Per-producer trial count is too low to beat chance for any
  meaningful statistic, even with real signal — fall back to the
  group-mean infoVal as the publishable number.

**Reporting recommendation.** Report **(i)** the group-mean
infoVal z and **(ii)** the proportion of producers whose
*individual* z clears 1.96 (with the face mask). Do not report only
the median per-producer z and treat values < 1.96 as evidence of
"no signal" — that conflates the per-producer noisiness of the
metric with the group-level question of whether the CI carries
information.

**Real example (Oliveira et al., 2019, Study 1, Trust condition).**
The same diagnostic recipe above, run on real published data. The
20 individual producers were each shown 300 trials (so the
trial-count-matched reference is simulated at 300):

```{r infoval-oliveira-2019, eval = FALSE}
# Per-producer infoVal
tc <- setNames(rep(300L, 20L), colnames(sm_trust))
iv <- infoval(sm_trust, noise_matrix, tc, iter = 1000L, seed = 1L)
#> per-producer: median z = 0.35, range [-1.69, 2.30], n above 1.96 = 2/20

# Group-mean CI infoVal (effective n_trials = 300 x 20 = 6000)
group  <- matrix(rowMeans(sm_trust), ncol = 1,
                 dimnames = list(NULL, "TRUST_group"))
iv_grp <- infoval(group, noise_matrix,
                  setNames(6000L, "TRUST_group"),
                  iter = 1000L, seed = 1L)
#> group-mean CI z = 11.39

# Face-mask lift on per-producer z
iv_masked <- infoval(sm_trust, noise_matrix, tc,
                     mask = face_mask(c(256L, 256L)),
                     iter = 1000L, seed = 1L)
#> masked median = 0.50  (vs unmasked median = 0.35; lift = +0.15)
```

This is exactly the pattern described in the troubleshooting list
above: per-producer z is centred around 0.35 with only 2 of 20
producers individually above the 1.96 threshold — and yet the
group-mean CI z is 11.39, *unmistakable* signal. The published
Trust/Friendly correlation of .69 (Table 2 of the paper) and the
observable trait dissociations in the group CIs (Figure 1 of the
paper) are both consistent with the group-level z, not with reading
the per-producer median as a "weak signal" verdict. Reporting
recommendation in practice: "the group-mean classification image
showed strong informational value (z = 11.39, n_eff = 6000); 2 of
20 producers individually exceeded z = 1.96."

> **Calibration note.** This output reflects the v0.2.1 fix in
> `simulate_reference_norms()` (sample stim ids without replacement
> when `n_trials <= n_pool`). v0.2.0 over-inflated the reference
> norm and gave `median z = -1.54` for the same data — a bug
> uncovered while replicating Oliveira et al. (2019), Study 1. If
> you computed infoVal under v0.2.0 and got systematically negative
> z-scores, re-run after upgrading.

## 7. Brief-RC end-to-end

Schmitz, Rougier & Yzerbyt (2024) introduced Brief-RC 12: each trial
shows 12 noisy faces (6 oriented, 6 inverted, drawn from a shared
noise pool); participant picks one. The recorded format is one row
per trial, columns `participant_id`, `trial`, `stimulus` (chosen pool
id), `response` (`+1` if oriented, `-1` if inverted). `rcicrely`'s
implementation is native (no rcicr `_brief` functions, which don't
exist in canonical rcicr v1.0.1).

```{r briefrc-demo, eval = FALSE}
# 1. Generate the noise pool (once per project) - heavy step.
rcicr::generateStimuli2IFC(
  base_face_files     = list(base = "data/base.jpg"),
  n_trials            = 600L,        # becomes the Brief-RC pool size
  img_size            = 256L,
  stimulus_path       = "data/stim",
  label               = "mystudy",
  ncores              = 1L,          # macOS-safe default
  return_as_dataframe = TRUE,        # capture the noise matrix!
  seed                = 1L
) |> saveRDS("data/noise_matrix.rds")  # cache for fast loading

# 2. Collect Brief-RC responses in the per-trial format.

# 3. Compute individual CIs.
res <- ci_from_responses_briefrc(
  responses       = my_brief_rc_responses,
  noise_matrix    = readRDS("data/noise_matrix.rds"),
  base_image_path = "data/base.jpg",
  method          = "briefrc12",
  scaling         = "matched"   # also rendered for plotting
)

# 4. Reliability assessment.
within  <- run_within(res$signal_matrix, seed = 1L)
between <- run_between(res_A$signal_matrix, res_B$signal_matrix,
                       seed = 1L)

# 5. Save rendered CI to PNG (visualization only - DO NOT use for stats)
if (requireNamespace("png", quietly = TRUE)) {
  for (j in seq_along(res$participants)) {
    img <- matrix(res$rendered_ci[, j], res$img_dims[1], res$img_dims[2])
    img <- pmin(pmax(img, 0), 1)
    png::writePNG(img, sprintf("plots/p%02d.png", j))
  }
}
```

**Brief-RC `infoVal`** ships in v0.2 as `infoval()` (see §6.5). The
reference distribution is matched to each producer's trial count
(not the pool size), which closes a calibration gap that biases
canonical 2IFC infoVal downward when applied to Brief-RC data. Pass
the **raw mask** (`res$signal_matrix`) — not `res$rendered_ci` or a
PNG.

## 8. Interpreting results and common pitfalls

**Sample size.** `rcicrely` aborts below 4 producers per condition
and warns below 30; a sample of N >= 60 producers is recommended for
stable reliability assessment. At N = 10 the metrics are computable
but their CIs are wide.

**Split-half.** Think of `r_hh` ~ 0.3 as weak, 0.5 as moderate, 0.7+
as strong producer-level consistency. `r_SB` is what to report when
asked about the full-sample CI.

**LOO flagging.** A flag is a starting point for investigation, not a
verdict. Use `flag_method = "mad"` if the SD rule is dominated by
the very outliers it tries to find.

**ICC(3,*) vs ICC(2,*).** Numerically close at high pixel counts but
ICC(3,*) is the correctly-specified model. Don't swap because of a
reviewer; report both with `variants = c("3_1", "3_k", "2_1", "2_k")`
if pressed.

**Cluster threshold != p-value.** `cluster_threshold = 2.0` is the
t-cutoff for forming clusters. Significance is decided by mass
relative to the null. The default is roughly two-tailed p < 0.05 at
large df, but at small N it under-rejects - inspect the t histogram.

**Raw vs. rendered CIs.** This package's reliability math is correct
**given** a raw signal matrix. Mode 1 (PNGs on disk) loads
`scaling(mask)` instead of `mask`; results will differ from what raw
responses would produce. The size of the difference depends on the
scaling that was applied at PNG-write time. Mode 2 is the safe path.

**`infoVal` correctness.** `rcicr::computeInfoVal2IFC()` extracts
the raw `$ci` from its input CI-list internally, so the canonical
2IFC path is safe regardless of `scaling`. The pitfall is in
hand-rolled `infoVal` implementations, for Brief-RC, or for any
custom code that consumes a bare matrix, which must be fed the
raw mask. Feeding rendered/scaled values silently distorts the
reference comparison and produces wrong significance verdicts.

## 9. References

- Brinkman, L., Todorov, A., & Dotsch, R. (2017). Visualising mental
  representations: A primer on noise-based reverse correlation in
  social psychology. *European Review of Social Psychology*, 28(1),
  333-361. <https://doi.org/10.1080/10463283.2017.1381469>
- Brinkman, L., Goffin, S., van de Schoot, R., van Haren, N. E. M.,
  Dotsch, R., & Aarts, H. (2019). Quantifying the informational
  value of classification images. *Behavior Research Methods*,
  51(5), 2059-2073. <https://doi.org/10.3758/s13428-019-01232-2>
- Cone, J., Brown-Iannuzzi, J. L., Lei, R., & Dotsch, R. (2021). Type I Error Is Inflated in the Two-Phase Reverse Correlation Procedure. *Social Psychological and Personality Science*, 12(5), 760-768. <https://doi.org/10.1177/1948550620938616>
- Dotsch, R. (2023). *rcicr*: Reverse-Correlation Image-Classification
  Toolbox. R package v1.0.1.
  <https://github.com/rdotsch/rcicr>
- Maris, E., & Oostenveld, R. (2007). Nonparametric statistical
  testing of EEG- and MEG-data. *Journal of Neuroscience Methods*,
  164(1), 177-190. <https://doi.org/10.1016/j.jneumeth.2007.03.024>
- McGraw, K. O., & Wong, S. P. (1996). Forming inferences about some
  intraclass correlation coefficients. *Psychological Methods*,
  1(1), 30-46. <https://doi.org/10.1037/1082-989X.1.1.30>
- Nichols, T. E., & Holmes, A. P. (2002). Nonparametric permutation
  tests for functional neuroimaging: A primer with examples.
  *Human Brain Mapping*, 15(1), 1-25.
  <https://doi.org/10.1002/hbm.1058>
- Oliveira, M., Garcia-Marques, T., Dotsch, R., & Garcia-Marques, L.
  (2019). Dominance and competence face to face: Dissociations
  obtained with a reverse correlation approach. *European Journal
  of Social Psychology*. <https://doi.org/10.1002/ejsp.2569>.
  Open data and analysis scripts: <https://doi.org/10.17605/osf.io/hr5pd>.
- Schmitz, M., Rougier, M., & Yzerbyt, V. (2024). Introducing the
  brief reverse correlation: An improved tool to assess visual
  representations. *European Journal of Social Psychology*.
  <https://doi.org/10.1002/ejsp.3100>
- Shrout, P. E., & Fleiss, J. L. (1979). Intraclass correlations:
  Uses in assessing rater reliability. *Psychological Bulletin*,
  86(2), 420-428. <https://doi.org/10.1037/0033-2909.86.2.420>

[rcicrdiagnostics-link]: https://github.com/olivethree/rcicrdiagnostics
