---
title: "rcicrely user guide"
author: "Manuel Oliveira"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{rcicrely user guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi       = 96
)
options(rcicrely.silence_scaling_warning = TRUE)  # quiet vignette
```

This guide answers two questions side by side: which function to call
for a given question, and why its output is trustworthy. Each function
gets a plain-language description, the signature, a worked example,
how to read the output, and the footguns to avoid.

## 1. What `rcicrely` is - and isn't

In reverse correlation (RC), participants' trial-level choices project
onto a pool of noise patterns to reconstruct the mental representation
they used. The output is a per-producer **classification image** (CI)
plus a group-level CI obtained by averaging.

`infoVal` (Brinkman et al., 2019) tells you whether a CI contains more
signal than chance. It does **not** tell you whether that signal is
*stable across producers*, *replicable on a different subset of
producers*, or *distinguishable from another condition's CI*.

`rcicrely` fills that gap with a pixel-level reliability toolkit:

- **Within-condition** (how consistent is a condition's CI across
  producers?): permuted split-half with Spearman-Brown,
  leave-one-out sensitivity, ICC(3,1) and ICC(3,k).
- **Between-condition** (is condition A's CI distinguishable from
  condition B's?): vectorised Welch t per pixel, cluster-based
  permutation with max-statistic FWER, bootstrap representational
  dissimilarity.

None of this requires a phase-2 trait-rating study. The metrics read
directly from the pixel-level signal produced by the original
producers, sidestepping the two-phase design Cone, Brown-Iannuzzi,
Lei, and Dotsch (2021) showed inflates Type I error.

### `infoVal` and `rcicrely` together

Since v0.2 `rcicrely` ships its own `infoval()` (see §6.5) that
handles **both 2IFC and Brief-RC** with a single function. The
reference distribution is keyed on each producer's actual trial
count rather than the full pool size (the calibration matters most
for Brief-RC, where producers see only a subset of the pool, but the
function works equivalently for 2IFC). The two checks remain
complementary: `infoval()` tests presence of signal; `rel_*()` tests
reliability and discriminability of signal.

`rcicr::computeInfoVal2IFC()` is still useful as an external cross-
check for 2IFC data. It pulls the raw `$ci` element from the rcicr
CI-list internally (`norm(matrix(target_ci[["ci"]]), "f")`), so the
scaling settings used at CI generation do not affect its output.

The scaling caveat applies when you compute `infoVal` outside that
path: both `rcicrely::infoval()` and any hand-rolled implementation
must be fed the raw mask, never a PNG-derived or otherwise scaled
matrix. §3 and §4 explain how each input mode delivers (or does
not deliver) the raw mask.

## 2. Installation

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("rdotsch/rcicr")      # 2IFC path only
remotes::install_github("olivethree/rcicrely")
```

If you only use Brief-RC you can skip `rcicr` - the Brief-RC CI
implementation is fully native (Schmitz, Rougier & Yzerbyt, 2024).

```{r load}
library(rcicrely)
```

## 3. Mental model: the signal matrix

Every reliability metric in this package takes a single object: a
**signal matrix**, pixels by participants. Get this right and the
rest of the package is a thin shell over it.

> **A note on the word *mask*.** The term has two distinct meanings
> in RC work and in this package:
>
> 1. **Noise mask** (often shortened to *mask* or *raw mask*): the
>    participant's *visual noise contribution* (the noise pattern
>    that, when superimposed on the base face, produces their
>    classification image). This is the core statistical object.
>    Every `rel_*()` function operates on a matrix of these.
> 2. **Region mask** (or *pixel mask*): a logical vector marking
>    which pixels to include in an analysis (oval face region, eyes
>    region, etc.). This is what `face_mask()` returns and what
>    `infoval()`'s `mask` argument consumes.
>
> Throughout this guide, "mask" without qualification means the noise
> mask. Pixel selection is always called "region mask" or referenced
> through `face_mask()`.

### 3.1 Two paths to the signal matrix

There are two ways to obtain the signal matrix, and the choice
matters because they produce mathematically different objects.

```
Mode 1: PNGs on disk           Mode 2: raw responses
          |                              |
   read_cis()                    ci_from_responses_2ifc() / _briefrc()
          |                              |
   extract_signal()                      |
          |                              |
   scaling(mask)                       mask                    <-- raw
          |                              |
          +---------> rel_*() <----------+
```

Mode 2 returns the **raw mask** (the participant's noise contribution
before any display transformation). Mode 1 cannot, because PNGs on
disk have always been rendered with a display scaling step baked in;
subtracting the base recovers `scaling(mask)`, not `mask`. §3.2
explains why this distinction is statistically load-bearing.

Prefer Mode 2 when raw responses are available. If you must use Mode
1, the package emits a one-time warning per session. To silence:

```{r silence-mode1, eval = FALSE}
# acknowledge per-call:
signal <- load_signal_matrix("data/cis/", "data/base.jpg",
                             acknowledge_scaling = TRUE)

# or silence for the whole session:
options(rcicrely.silence_scaling_warning = TRUE)
```

If you generated the PNGs yourself with
`rcicr::generateCI2IFC(scaling = "none")`, the rendered output is
effectively raw and the warning is a false positive.

### 3.2 Raw mask vs. rendered CI

A CI exists in two forms.

A **raw mask** is the participant's noise contribution before any
display transformation. It is the object every reliability metric
operates on. For Brief-RC it is
`(noise_pool[, chosen_ids] %*% signed_responses) / n_unique`
(Schmitz's `genMask()`); for 2IFC it is rcicr's `$ci` element. Values
are typically in a narrow band around zero (roughly +/- 0.05).

The **rendered CI** is what gets saved to PNG. It is
`base + scaling(mask)`, where `scaling()` may stretch the mask to the
base image's range ("matched"), apply a constant gain, or do nothing.
Scaling is a *display* transformation, not a statistical one.
Subtracting the base from a rendered CI does **not** recover the raw
mask:

```
PNG_pixels - base = scaling(mask)   (NOT  mask)
```

The metrics introduced in §5 and §6 react differently to scaling:

| Metric                                       | Behavior under scaling                                       |
|---------------------------------------------:|:-------------------------------------------------------------|
| `rel_split_half`, `rel_loo`                  | Survives uniform scaling. Breaks under per-CI "matched".     |
| `rel_dissimilarity` (Pearson r half)         | Same.                                                        |
| `rel_dissimilarity` (Euclidean half)         | **Distorted** by any scaling.                                |
| `rel_icc`                                    | **Distorted** by any scaling.                                |
| `pixel_t_test`, `rel_cluster_test`           | **Distorted** by per-CI scaling.                             |
| `rcicr::computeInfoVal2IFC()` (standard)     | Uses raw `$ci` internally, unaffected.                       |
| Hand-rolled `infoVal` (Brief-RC, custom)     | **Distorted** by any scaling. Use the raw mask.              |

### 3.3 A concrete demonstration

Here is what scaling does to the same underlying mask.

```{r scaling-demo, fig.height = 3.5, fig.width = 7}
set.seed(1)
n_pix <- 32L * 32L
n_p   <- 20L
mask  <- matrix(rnorm(n_pix * n_p, sd = 0.02), n_pix, n_p)

# Render: stretch each column to [-0.5, 0.5] (rcicr's "matched" /
# auto-scale option; the rcicr default for visualisation)
rendered <- apply(mask, 2L, function(m) {
  rng <- diff(range(m)); if (rng == 0) m else m * (1 / rng)
})

c(
  raw_max_range      = round(diff(range(mask)),     3L),
  rendered_max_range = round(diff(range(rendered)), 3L)
)

# Pearson r between same-column raw and rendered: ~ 1
# (a perfect linear stretch preserves Pearson)
mean(vapply(seq_len(n_p),
            function(j) cor(mask[, j], rendered[, j]),
            numeric(1L)))

# But the variance ratios that drive ICC are very different
var(rowMeans(mask))      # variance of the group mean over pixels
var(rowMeans(rendered))  # same metric, on the rendered version
```

Pearson r per column is exactly 1 (linear scaling is correlation-
preserving), but the row-mean variance the ICC formula depends on is
~25x larger on the rendered version. ICC(3,*) computed on rendered
data is therefore not the same ICC computed on the raw mask. The
same logic applies to Euclidean distance and to `infoVal`.

### 3.4 Base image preparation

Both 2IFC and Brief-RC pipelines depend on a base face image. Both
the visible quality of generated CIs and the reliability statistics
that follow are sensitive to how this image is prepared.
Requirements:

- **Square** (e.g. 256x256, 512x512).
- **Grayscale** (single channel). RGB images error out of rcicr
  with non-conformable-array messages.
- **Pixel range `[0, 1]`** (the convention `png::readPNG` and
  `jpeg::readJPEG` produce).
- **Centred face** with eye / nose / mouth roughly at the geometry
  the Schmitz oval mask assumes (eyes near the upper third, mouth
  near the lower third). The
  built-in `face_mask(region = ...)` geometries assume this layout;
  custom or off-centre faces need their own mask geometry.

The conventional base face is a *morph* of several individual faces,
which removes idiosyncratic features and centres the geometry.
[webmorphR](https://github.com/debruine/webmorphR) (DeBruine, 2022)
is the current best-in-class tool for producing one. Typical
pipeline:

```{r webmorphR-pipeline, eval = FALSE}
# install.packages("webmorphR")
library(webmorphR)

stim <- read_stim("path/to/raw_face_images/") |>
  auto_delin() |>                       # auto landmark delineation
  align(procrustes = TRUE) |>           # Procrustes alignment
  crop(width = 0.85, height = 0.85) |>  # tight crop around the face
  to_size(c(256, 256)) |>               # resize to rcicr-friendly size
  greyscale() |>                        # single channel, in [0, 1]
  avg()                                 # morph into one average face

write_stim(stim, dir = "stimuli/", names = "base", format = "png")
```

`write_stim()` produces `stimuli/base.png`, ready for
`rcicr::generateStimuli2IFC(base_face_files = list(base = "stimuli/base.png"))`.
For non-square crops, oval-region operations, and feature-aligned
masks (which then plug into [`load_face_mask()`](#section-load-face-mask)),
see `?webmorphR::mask_oval` and `?webmorphR::crop_tem`. Cite
DeBruine (2022).

If you do not have a stack to morph (e.g. you are working from a
single base photograph), prepare it by hand: any image editor will
do (GIMP / Krita / Photoshop / PowerPoint export). Convert to
grayscale, crop to a square that fits the face within the central
~70%, resize to the target dimension, save as PNG.

### 3.5 A recurring real-world example

The synthetic demos in the function-reference sections below are
deliberately small so the vignette knits in seconds. To see what
each function looks like on a published RC dataset, the sections
that follow each include a short **Real example** callout drawn
from one source:

> Oliveira, M., Garcia-Marques, T., Dotsch, R., & Garcia-Marques,
> L. (2019). Dominance and competence face to face: Dissociations
> obtained with a reverse correlation approach. *European Journal
> of Social Psychology*. <https://doi.org/10.1002/ejsp.2569>

In Study 1, 200 participants completed a 2IFC reverse-correlation
task with **300 trials each** on a 256x256 male base face,
across 10 trait conditions in a between-subjects design (20
producers per trait). Trait labels: `Dominant`, `Submissive`,
`Trust`, `Untrust`, `Friendly`, `Unfriendly`, `Intelligent`,
`Unintelligent`, `Competent`, `Incompetent`. The full data,
parameter file, and analysis scripts are openly available on OSF:
<https://doi.org/10.17605/osf.io/hr5pd>.

The "Real example" callouts below show the call you would make
against this data and the output captured from a complete replication
run. The chunks are marked `eval = FALSE` because the data is not
bundled with the package; every output shown is from the published
dataset.

## 4. Function reference: I/O

The six functions in this section get your data into the package's
standard signal-matrix shape (`pixels x participants`). Two of them
(`ci_from_responses_2ifc()` and `ci_from_responses_briefrc()`) take
**trial-level response data**; before diving into them the shape and
coding conventions for that data deserve a quick callout.

### Input response-data shape

One row per trial. The format can be a `data.frame`, `data.table`,
`tibble`, or anything that behaves like a data frame; the source
(CSV, RData, Parquet, programmatic construction) does not matter,
only that the columns below are present.

| Column           | Purpose                                                                                 |
|------------------|-----------------------------------------------------------------------------------------|
| `participant_id` | Identifier per producer. Character or integer.                                          |
| `stimulus`       | Stimulus / pool id. Numbering must match what was used at stimulus generation.          |
| `response`       | `+1` or `-1` (see below).                                                               |
| `rt` (optional)  | Response time in milliseconds. Not consumed by `rcicrely`; useful for `rcicrdiagnostics`. |

**Response coding.**

- **2IFC**: `response = +1` if the participant chose the *oriented*
  variant on that trial, `-1` if the *inverted* variant. Each trial
  has its own unique stimulus pair, so `stimulus` indexes the *pair*
  and ranges `1:n_trials`.
- **Brief-RC 12** (Schmitz, Rougier & Yzerbyt, 2024): on each trial
  the participant sees 12 noisy faces (6 oriented + 6 inverted from
  6 distinct underlying noise patterns) and picks one. Recorded as
  one row per trial, with `stimulus` = pool id of the chosen
  pattern and `response = +1` if the oriented variant was chosen,
  `-1` if the inverted. The same `stimulus` id can repeat across
  trials.

> **Watch out for the `{0, 1}` miscoding.** This is the single most
> common silent failure in RC pipelines: experiment software
> sometimes records "left" / "right" or "first" / "second" as
> `0` / `1`, and the analyst forgets to recode. Pass `{0, 1}`
> responses to `ci_from_responses_*()` and the resulting CI is
> approximately blank (the math leans heavily on the sign). Both
> functions error out at the input boundary if responses are not in
> `{-1, +1}`. If you see that error, the fix is one line:
> `responses$response <- 2 * responses$response - 1`.

> **Brief-RC: do not record the 11 unselected faces.** Each Brief-RC
> trial shows 12 faces but the participant chooses one. Record only
> the chosen face (one row per trial, as above). Do *not* record
> the 11 unselected faces as additional rows with `response = 0`,
> and do *not* use the "expanded" 12-rows-per-trial format that
> weights chosen as `+1` and unchosen as `-1/11`. Neither
> convention matches the Schmitz `genMask()` formula
> (`mask = noise[, chosen] %*% chosen_responses / n_unique_chosen`)
> that `ci_from_responses_briefrc()` implements. Unselected faces'
> noise patterns simply do not enter the multiplication; their
> absence *is* the zero.

### 4.1 `read_cis()`

**What it does.** Reads every PNG/JPEG in a directory, converts each
to grayscale, returns a pixels by participants numeric matrix.

**When to use it.** You have one CI image per producer already saved
on disk and you want to load them into R as a numeric matrix. Pair
with [`extract_signal()`](#section-extract-signal) to get a
base-subtracted signal matrix.

```{r read-cis-demo}
# Make a tiny synthetic dataset on disk to demonstrate against.
demo_dir <- tempfile("ci_demo_"); dir.create(demo_dir)
base_img <- matrix(0.5, 32L, 32L)
png_path <- function(i) file.path(demo_dir, sprintf("p%02d.png", i))
if (requireNamespace("png", quietly = TRUE)) {
  for (i in 1:6) {
    img <- pmin(pmax(base_img + matrix(rnorm(32 * 32, sd = 0.05), 32, 32),
                     0), 1)
    png::writePNG(img, png_path(i))
  }
}
```

```{r read-cis-call, eval = requireNamespace("png", quietly = TRUE)}
cis <- read_cis(demo_dir, acknowledge_scaling = TRUE)
dim(cis)
attr(cis, "img_dims")
```

**Reading the result.** `nrow = n_pixels`, `ncol = n_files`. The
`img_dims` attribute carries `c(nrow, ncol)` of the original image so
plotting helpers know how to reshape.

**Common mistakes.** Forgetting to subtract the base before passing
to a `rel_*` function (use [`extract_signal()`](#section-extract-signal)
or the convenience wrapper [`load_signal_matrix()`](#section-load-signal-matrix)).
Mixing CIs of different sizes - aborts loudly. Trusting reliability
numbers without reading the raw vs. rendered note (chapter 3).

### 4.2 `extract_signal()` {#section-extract-signal}

**What it does.** Subtracts the base image from each column of a CI
matrix, so reliability metrics see only the participant-specific
contribution.

**When to use it.** You have the output of [`read_cis()`](#read_cis)
and you want a base-subtracted signal matrix.

```{r extract-signal-demo, eval = requireNamespace("png", quietly = TRUE)}
base_path <- file.path(demo_dir, "_base.png")
png::writePNG(base_img, base_path)
signal <- extract_signal(cis, base_image_path = base_path,
                         acknowledge_scaling = TRUE)
dim(signal); attr(signal, "img_dims")
```

**Reading the result.** Same shape as `cis`, with the `img_dims`
attribute preserved. Column names propagate.

**Common mistakes.** Pointing at the wrong base (a generated CI PNG
of the right size will pass the dimension check). Treating the result
as the raw mask; it is `scaling(mask)` if PNGs were rendered with any
scaling. See chapter 3.

### 4.3 `load_signal_matrix()` {#section-load-signal-matrix}

**What it does.** Convenience wrapper that calls [`read_cis()`](#read_cis)
and [`extract_signal()`](#section-extract-signal) in one go.

**When to use it.** Most of the time, when you don't need to do
anything between reading and subtracting (e.g. masking pixels,
swapping the base).

```{r load-signal-demo, eval = requireNamespace("png", quietly = TRUE)}
signal <- load_signal_matrix(
  dir                 = demo_dir,
  base_image_path     = base_path,
  pattern             = "^p[0-9]+\\.png$",  # exclude _base.png
  acknowledge_scaling = TRUE
)
dim(signal)
```

### 4.4 `read_noise_matrix()`

**What it does.** Loads the pixels by trials noise pool used at
stimulus generation. Three formats: rcicr `.Rdata` (reconstructs from
`stimuli_params` + `p`), saved `.rds`, or whitespace-delimited text.

**When to use it.** You are computing Brief-RC CIs and want to load
the noise pool once, then reuse it across multiple
[`ci_from_responses_briefrc()`](#section-cifr-briefrc) calls.

```{r read-noise-demo, eval = FALSE}
# from the rcicr rdata (slow - reconstructs each trial via
# rcicr::generateNoiseImage):
noise <- read_noise_matrix("data/rcicr_stimuli.Rdata")

# from a cached rds you saved from the rcicr return value (fast):
noise <- read_noise_matrix("data/noise_matrix.rds")
```

**Common mistakes.** Assuming the rcicr `.Rdata` contains a `stimuli`
object (it does not - the pixels-by-trials matrix is the *return
value* of `generateStimuli2IFC(return_as_dataframe = TRUE)`, never
saved to the file). Cache to `.rds` once and load from there.
Multiple stale `.rdata` files in one directory: rcicr timestamps
filenames; pick the most recent by mtime if you have several.

#### What is in an rcicr `.RData`

`load("rcic_stimuli.Rdata")` adds the following objects to your
environment. Names are short and not self-explanatory; the table
below is the field guide.

| Object                | Plain-language description                                                                                                                                                                                                                                                              |
|-----------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| `base_face_files`     | Named list of file paths to the original face images. The list names (e.g. `"base"`) are the labels you pass downstream as `baseimage = "..."`.                                                                                                                                          |
| `base_faces`          | Base images themselves, loaded as numeric matrices of grayscale pixels in `[0, 1]`. One matrix per label; shape `img_size x img_size`.                                                                                                                                                  |
| `img_size`            | Side length of the (square) images in pixels.                                                                                                                                                                                                                                            |
| `n_trials`            | Number of stimulus pairs (and unique noise patterns) generated.                                                                                                                                                                                                                          |
| `noise_type`          | Type of noise basis. The rcicr default and only well-tested option is `"sinusoid"`.                                                                                                                                                                                                      |
| `p`                   | The **noise basis**. A list with `$patches` (a stack of standard sinusoidal patterns at multiple scales, orientations, phases) and `$patchIdx` (an index that maps positions in a parameter vector to entries in `$patches`). Think of `$patches` as a dictionary of sinusoidal "ingredients". |
| `stimuli_params`      | The **per-trial recipe** for combining `p`'s ingredients. A named list of matrices, one per base face. Each row is one trial; each entry is a contrast weight. `rcicr::generateNoiseImage(stimuli_params[[base]][i, ], p)` reconstructs trial `i`.                                       |
| `seed`                | Random seed used at generation, for reproducibility.                                                                                                                                                                                                                                      |
| `label`, `stimulus_path`, `trial`, `generator_version`, `use_same_parameters` | Bookkeeping; not consumed by analysis.                                                                                                                                                                                          |
| `reference_norms`     | A vector of random-responder Frobenius norms. **Not present at first.** Created and inserted in place the first time `rcicr::computeInfoVal2IFC()` is called on the file. Copy the rdata first if you want it untouched.                                                                |

The load-bearing objects for analysis are `base_faces`,
`stimuli_params`, `p`, and `img_size`. The full set of trial-level
noise images is **not stored**; it is recomputed on demand via
`generateNoiseImage()` (which `read_noise_matrix()` calls in a
loop). For 256x256 images and `n_trials = 770` (rcicr default) this
takes a few seconds; cache the result with `saveRDS()` to skip it
on subsequent runs.

> **Filename note.** On macOS the file is saved with a lowercase
> `.Rdata` extension. `list.files(pattern = "\\.RData$")` will miss
> it; use `ignore.case = TRUE`.

### 4.5 `ci_from_responses_2ifc()` {#section-cifr-2ifc}

**What it does.** Wraps `rcicr::batchGenerateCI2IFC()` with all known
gotchas pre-handled, returns a uniform result list.

**When to use it.** You have 2IFC trial-level responses and want
per-producer CIs in one call. Requires `rcicr` installed.

```{r cifr-2ifc-demo, eval = FALSE}
res <- ci_from_responses_2ifc(
  responses       = my_responses,
  rdata_path      = "data/rcicr_stimuli.Rdata",
  baseimage       = "base",
  participant_col = "participant_id",
  keep_rendered   = FALSE   # set TRUE if you want $rendered_ci for plotting
)
signal <- res$signal_matrix
```

**Reading the result.** `$signal_matrix` is the **raw mask** (rcicr's
`$ci` per producer) - feed this to `rel_*`. `$participants` and
`$img_dims` are convenience metadata. `$rcicr_result` is the raw
rcicr return value if you want anything else. `$rendered_ci`
(present only if `keep_rendered = TRUE`) is `base + scaling(mask)`
per producer - **visualization only**, do not feed to `rel_*`.

**Common mistakes.** Inventing a `base_image_path` argument - the base
is read from the rdata via `baseimage` *label*, not a path. Passing
`$rendered_ci` to `rel_*`. Trusting `scaling = "autoscale"` to be
"raw" - the rendered output is scaled even when the signal_matrix
isn't.

### 4.6 `ci_from_responses_briefrc()` {#section-cifr-briefrc}

**What it does.** Native Brief-RC 12 implementation of Schmitz's
`genMask()` formula. No `rcicr::*_brief` calls (those don't exist in
the upstream rcicr v1.0.1).

**When to use it.** You have Brief-RC 12 trial-level responses and
the noise pool used at stimulus generation.

```{r cifr-briefrc-demo, eval = FALSE}
res <- ci_from_responses_briefrc(
  responses       = my_responses,    # one row per trial
  rdata_path      = "data/rcicr_stimuli.Rdata",
  base_image_path = "data/base.jpg",
  method          = "briefrc12",
  scaling         = "none"           # raw mask only - no $rendered_ci
)
signal <- res$signal_matrix
```

To also get the rendered CI for plotting:

```{r cifr-briefrc-rendered, eval = FALSE}
res <- ci_from_responses_briefrc(
  responses       = my_responses,
  noise_matrix    = my_noise,
  base_image_path = "data/base.jpg",
  scaling         = "matched"        # rcicr auto-scale (used for
                                     # visualisation by Schmitz et al.
                                     # 2024 in Experiment 2)
)
# res$signal_matrix is the raw mask, feed to rel_*
# res$rendered_ci is base + matched(mask), for PNG / plotting only
```

**Reading the result.** `$signal_matrix` is **always** the raw mask,
regardless of `scaling`. `$rendered_ci` exists only when
`scaling != "none"`. `$participants`, `$img_dims`, `$scaling` are
metadata.

**Common mistakes.** Passing the "expanded" 12-rows-per-trial format -
Brief-RC 12 is one row per trial (the chosen face), not 12. Asking
for `method = "briefrc20"` (deferred). Using `$rendered_ci` for
downstream stats.

```{r cleanup-demo, include = FALSE}
unlink(demo_dir, recursive = TRUE)
```

### 4.7 `load_face_mask()` {#section-load-face-mask}

**What it does.** Reads a binary mask image from disk (PNG or JPEG)
and returns a logical vector of length `prod(img_dims)` in
column-major order (the format the rest of the package's mask
arguments expect). Companion to [`face_mask()`](#section-face-mask)
(parametric / programmatic) for users who want to work with
hand-painted, webmorphR-generated, or otherwise externally-produced
masks.

**When to use it.** You have a binary mask image (e.g.,
`webmorphR::mask_oval()` output saved with `write_stim()`, a
hand-painted GIMP / PowerPoint mask, or any other source). You want
to apply it via `infoval()`'s `mask` argument, via
`plot_agreement_map()`'s `mask` argument, or via row-subsetting on a
signal matrix.

```{r load-face-mask-demo, eval = FALSE}
fm <- load_face_mask("masks/oval_256.png",
                     threshold     = 0.5,
                     expected_dims = c(256L, 256L))
mean(fm)               # fraction of pixels included
infoval(signal_matrix, noise_matrix, trial_counts,
        mask = fm, iter = 1000L, seed = 1L)
```

**Reading the result.** A logical vector. White / light pixels
(luminance above `threshold`, default 0.5) become `TRUE`; dark
pixels become `FALSE`. RGB images are converted to luminance via
ITU-R BT.709 weights.

**Common mistakes.** Mask dimensions different from the analysis
image: pass `expected_dims = c(nrow, ncol)` to catch this at the
input boundary instead of having the math go silently wrong
downstream. Mask saved with the *opposite* convention (black =
include, white = exclude): the easy fix is `!load_face_mask(...)` to
invert the logical vector.

## 5. Function reference: within-condition reliability

For the rest of this guide we'll work with a small synthetic dataset
where two conditions have different latent signals.

```{r fake-data}
set.seed(1)
make_condition <- function(region_top, n_pix = 32L * 32L,
                           n_p = 30L, snr = 0.3) {
  noise <- matrix(rnorm(n_pix * n_p, sd = 0.05), n_pix, n_p)
  mask <- if (region_top) {
    c(rep(1, n_pix / 2L), rep(0, n_pix / 2L))
  } else {
    c(rep(0, n_pix / 2L), rep(1, n_pix / 2L))
  }
  out <- noise + snr * outer(mask, runif(n_p, 0.7, 1.3))
  attr(out, "img_dims") <- c(32L, 32L)
  out
}
sig_A <- make_condition(region_top = TRUE,  n_p = 30L)
sig_B <- make_condition(region_top = FALSE, n_p = 30L)
```

### 5.1 `rel_split_half()` {#section-split-half}

**What it does.** Estimates how stable a condition's group CI is
across random halves of producers. Reports the per-permutation
correlation between halves and the Spearman-Brown projection to the
full sample.

**When to use it.** You want to answer "would I get the same group
CI if I'd run a different half of my participants?". `r_SB` is
usually the headline number.

```{r rel-split-half}
r_sh <- rel_split_half(sig_A, n_permutations = 500L, seed = 1L,
                       progress = FALSE)
print(r_sh)
```

```{r plot-split-half, fig.cap = "Permutation distribution of split-half r."}
plot(r_sh)
```

**Reading the result.** `$r_hh` = mean per-split r. `$r_sb` =
Spearman-Brown projected reliability of the full sample (report this
unless asked otherwise). `$ci_95`, `$ci_95_sb` are percentile CIs.
`$distribution` is the full per-permutation r vector.

**Common mistakes.** Reporting `$r_hh` instead of `$r_sb` when the
audience cares about the full-sample CI. Running with
`n_permutations < 200` and trusting the CIs.

**Real example (Oliveira et al., 2019, Study 1).** Per-trait
group-CI reliability for the Trust and Dominant conditions (20
producers each, 300 trials, 256x256 image):

```{r split-half-oliveira-2019, eval = FALSE}
# sm_trust, sm_dominant: pixels x 20 producer signal matrices
# (built from the response data; see study1_replication.Rmd)
rel_split_half(sm_trust,    n_permutations = 500L, seed = 1L)
#> Trust:    r_hh = 0.405, r_sb = 0.577, 95% CI on r_sb = [0.539, 0.616]
rel_split_half(sm_dominant, n_permutations = 500L, seed = 1L)
#> Dominant: r_hh = 0.262, r_sb = 0.415, 95% CI on r_sb = [0.380, 0.450]
```

The Trust CI replicates with a clearly higher group-level
reliability than the Dominant CI, consistent with the inter-rater
agreement ordering reported in Table 1 of Oliveira et al. (2019).

### 5.2 `rel_loo()` - influence screening (not a reliability metric) {#section-rel-loo}

**What it does.** For each producer, correlates the full-sample
group CI with the group CI computed without that producer, and
returns a **z-scored** version of those correlations. Producers
whose z-score sits clearly below zero have a disproportionate
influence on the group pattern and are flagged for inspection.

**When to use it.** To screen for producers whose individual CI is
far from the group pattern - often a sign of response miscoding,
fatigue, or a genuinely atypical mental representation. This is a
diagnostic; it is *not* a reliability statistic (see the note at the
end of this section).

```{r rel-loo}
r_loo <- rel_loo(sig_A, flag_threshold = 2.5, flag_method = "sd")
print(r_loo)
```

The default is mean +/- SD; use `flag_method = "mad"` for a
median +/- MAD rule that doesn't get pulled around by the very
outliers it tries to detect:

```{r rel-loo-mad}
r_loo_mad <- rel_loo(sig_A, flag_threshold = 2.5, flag_method = "mad")
c(sd_threshold = r_loo$threshold, mad_threshold = r_loo_mad$threshold)
```

```{r plot-loo, fig.height = 5, fig.cap = "z-scored LOO influence per producer."}
plot(r_loo)
```

For a tidy ordered table of producers by influence, use
`rel_loo_z()`:

```{r rel-loo-z, eval = FALSE}
head(rel_loo_z(r_loo))
#   participant_id  correlation  z_score  flag
# 1     outlier01        0.9714    -3.12  TRUE
# 2           p07        0.9942    -0.84  FALSE
# ...
```

**Reading the result.** `$z_scores` is the informative quantity -
plot / report it directly. `$correlations` is retained for
transparency (values near 1 by construction because the full-sample
mean and the leave-one-out mean share `(N - 1) / N` of their data).
`$flagged` lists ids below threshold. `$summary_df` is a tidy table
sorted by `z_score`.

**Why you should not read `$correlations` as reliability.** `r_loo`
values are almost always in `[0.95, 0.999]` at N = 30, even on very
noisy CIs, because the two vectors being correlated overlap in
`(N - 1) / N` of their data. A naive reader can misinterpret
"LOO r = 0.98" as "the CIs are 98% reliable"; this is wrong. Report
reliability via `rel_split_half()` or `rel_icc()` (§5.1, §5.3).

**Common mistakes.** Treating `$flagged` as "drop these producers" -
investigate first. A flagged producer may have a genuinely atypical
mental representation rather than bad data. Cross-check with
`rcicrdiagnostics` to rule out response-coding errors.

**Real example (Oliveira et al., 2019, Study 1, Trust condition).**
Median raw `r_loo` is .992 (MAD = .001), illustrating the near-1
absolute value structurally. Zero producers cleared the default
MAD-based flag threshold, but the z-scored ordering is still
informative for inspection priority:

```{r loo-oliveira-2019, eval = FALSE}
loo <- rel_loo(sm_trust, flag_method = "mad")
head(rel_loo_z(loo), 3)
#>   participant_id  correlation  z_score  flag
#> 1          18043       0.9905    -1.68  FALSE
#> 2          18051       0.9911    -0.99  FALSE
#> 3          18131       0.9913    -0.79  FALSE
```

The producers at the top of the z-ordering are the ones whose
removal moves the group CI most relative to the rest. Even when
none clear the flag threshold, this is the inspection order.

### 5.3 `rel_icc()` {#section-rel-icc}

**What it does.** Two-way mixed ICC via direct mean squares. Default
returns ICC(3,1) and ICC(3,k). Pixels are fixed (the image grid),
participants are random (drawn from a producer population).

**When to use it.** Single-scalar reliability summary. ICC(3,k) is
"how stable is the group-mean CI?", ICC(3,1) is "how informative is
one producer's CI?".

```{r rel-icc}
r_icc <- rel_icc(sig_A, variants = c("3_1", "3_k"))
print(r_icc)
```

To also report ICC(2,*):

```{r rel-icc-both}
r_icc_full <- rel_icc(sig_A, variants = c("3_1", "3_k", "2_1", "2_k"))
print(r_icc_full)
```

**Reading the result.** `$icc_3_1`, `$icc_3_k`, `$icc_2_1`,
`$icc_2_k` (the latter two NA unless requested). `$ms_rows`,
`$ms_cols`, `$ms_error` are the underlying ANOVA mean squares for
transparency. `$model` is a description string the print method shows.

**Common mistakes.** Reporting ICC(2,*) without need. ICC(2,*) treats
pixels as a random sample from a pixel population; in RC studies
pixels are a fixed image grid, so ICC(3,*) is the correctly-specified
model. The numbers are usually close at high pixel counts, but
ICC(3,*) is the defensible choice. Use
`variants = c("3_1", "3_k", "2_1", "2_k")` when both are needed.

ICC is **strongly** scale-sensitive. Use the raw mask only.

**Real example (Oliveira et al., 2019, Study 1).** Two-way mixed
ICC on 65,536-pixel raw masks, 20 producers per condition:

```{r icc-oliveira-2019, eval = FALSE}
rel_icc(sm_trust)
#> ICC(3,1) = 0.1008,  ICC(3,k) = 0.6917
rel_icc(sm_dominant)
#> ICC(3,1) = 0.0489,  ICC(3,k) = 0.5068
```

ICC(3,1) (the resolution-comparable single-producer reliability)
is clearly higher for Trust than for Dominant, matching the
reliability ordering reported in the original paper. ICC(3,k) is
high in absolute terms but the resolution asymptote is a known
property of ICC(*,k) at large pixel counts; report ICC(3,1) as the
primary cross-study comparable number.

### 5.4 `run_within()` {#section-run-within}

**What it does.** Calls all three within-condition metrics and bundles
the result for joint printing/plotting.

```{r run-within}
within_A <- run_within(
  sig_A,
  n_permutations = 500L,
  seed           = 1L,
  progress       = FALSE
)
print(within_A)
```

`$results$split_half` and `$results$icc` are the two reliability
result objects. `plot(within_A)` lays them out side by side.
Leave-one-out influence screening is run separately with
`rel_loo()` (it is an influence diagnostic, not a reliability
statistic; see §5.2).

## 6. Function reference: between-condition inference

The two examples in §5 (`sig_A`, `sig_B`) use a half-image signal,
which produces *very* strong contrasts (useful for demonstrating
that within-condition reliability metrics work, but the
between-condition cluster figures end up looking like two solid blobs
rather than realistic localized clusters). For §6.1–§6.2 we therefore use a
small additional fixture with **localized circular signal**: two
small "patches" at different positions on a 32 x 32 grid, mimicking
how real RC signal tends to cluster around face features rather than
fill the whole image.

```{r fake-data-localized}
set.seed(7)
make_localized <- function(centre, n_p = 30L, n_side = 32L,
                           snr = 0.20, radius = 0.18, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)
  rr <- row(matrix(0, n_side, n_side)) / n_side - centre[1]
  cc <- col(matrix(0, n_side, n_side)) / n_side - centre[2]
  d  <- sqrt(rr^2 + cc^2)
  m  <- as.vector(pmax(0, 1 - d / radius))
  out <- snr * outer(m, runif(n_p, 0.7, 1.3)) +
         matrix(rnorm(n_side^2 * n_p, sd = 0.05), n_side^2, n_p)
  attr(out, "img_dims") <- c(n_side, n_side)
  out
}
# Condition A: signal in upper-left; Condition B: signal in lower-right
sig_A_loc <- make_localized(c(0.30, 0.30), n_p = 30L, seed = 11L)
sig_B_loc <- make_localized(c(0.70, 0.70), n_p = 30L, seed = 12L)
```

### 6.1 `pixel_t_test()`: building block, not standalone test {#section-pixel-t}

**What it does.** Computes a Welch's t-statistic per pixel between
two condition signal matrices. Welch (unequal variances) is the
correct choice because conditions can differ in both N and
per-pixel variance. With `paired = TRUE`, computes a paired t on
the per-producer difference instead.

**Why this function exists.** 65,000 simultaneous t-tests are
statistically meaningless on their own; this is a *building block*,
not an inferential test. Three legitimate uses:

1. **Internal use by `rel_cluster_test()`**: every permutation of
   the cluster test recomputes a t-map; `pixel_t_test()` exposes
   that computation.
2. **Diagnostic visualisation**: inspecting the raw t-map *before*
   running the cluster test helps you pick a sensible cluster-
   forming threshold and check whether any spatially-coherent
   structure is even there to test.
3. **Custom inference pipelines**: implementing TFCE by hand,
   defining non-standard cluster geometries, or applying a
   different multiple-comparison correction (FDR, max-T, etc.).

**Do not** threshold the returned vector at `|t| > 1.96` and report
"significant pixels". At 65,536 pixels you would get ~3,300 false
positives by chance under H0 (any α), and spatial clustering in
real CI data makes that worse. Use `rel_cluster_test()` for
inference.

```{r pixel-t}
t_vec <- pixel_t_test(sig_A_loc, sig_B_loc)
range(t_vec)
```

A quick visualization makes the output structure obvious. Positive
(red) regions indicate where condition A's signal is stronger than
B's; negative (blue) regions where B's is stronger than A's; near-
white pixels carry no consistent difference. **This is descriptive
only** (there is no inference yet):

```{r pixel-t-viz, fig.cap = "Raw pixel-wise Welch t-map between two conditions with localized signal patches. Positive (red) = condition A larger; negative (blue) = condition B larger. Note the two distinct hot regions corresponding to the injected signal locations; the rest of the image is dominated by noise. This map is descriptive only (significance comes from rel_cluster_test())."}
local({
  op <- par(mar = c(1, 1, 3, 6) + 0.1)
  on.exit(par(op))
  n_side <- 32L
  tmap   <- matrix(t_vec, n_side, n_side)
  rng    <- max(abs(t_vec))
  graphics::image(
    seq_len(n_side), seq_len(n_side),
    t(tmap[n_side:1L, ]),
    col  = grDevices::hcl.colors(256L, "RdBu", rev = TRUE),
    zlim = c(-rng, rng),
    main = "Raw pixel-wise t-map (descriptive only)",
    axes = FALSE, xlab = "", ylab = "",
    asp  = 1, useRaster = TRUE
  )
  graphics::box(col = "grey80", lwd = 0.5)
})
```

The hot blue/red regions are exactly where the two conditions
differ; everything else is noise. To turn this descriptive map into
a statement about *which* differences are reliable, we move on to
the cluster test.

### 6.2 `rel_cluster_test()`: spatially-localized inference with FWER control {#section-cluster}

**What it does.** Identifies *spatially contiguous regions* of the
image where two conditions differ, with family-wise error control
across the whole image (so you do not over-claim significance from
the kind of mass-testing problem §6.1 warned about).

**The vocabulary.** Three terms are unavoidable; here is what each
one means in plain language.

- **Cluster-forming threshold** (`cluster_threshold`, default
  `2.0`). A t-value cutoff used at the *first* step: any pixel with
  `|t| > 2.0` is "in", everything else is "out". This is *not* a
  significance threshold; it is the cutoff for deciding which
  pixels are even candidates for being part of a cluster.
  Significance is decided afterwards. Lower thresholds catch more
  diffuse signal; higher thresholds isolate sharper peaks.
- **4-connectivity**. Once we have a binary "in / out" image, we
  group adjacent "in" pixels into clusters. *4-connectivity*
  treats two pixels as adjacent only if they share an edge
  (up / down / left / right), i.e. the four orthogonal neighbours.
  *8-connectivity*, the alternative, also counts diagonals.
  4-connectivity is the more conservative choice: a one-pixel-wide
  diagonal stripe is split into isolated pixels and forms no
  cluster, whereas 8-connectivity would chain them. Suits face-CI
  signal, which tends to be contiguous in orthogonal directions.

  ```
  4-connectivity:    8-connectivity:
       . X .              X X X
       X o X              X o X
       . X .              X X X
   (o is the centre, X are neighbours, . are non-neighbours)
  ```

- **Cluster mass** (`mass`). For each candidate cluster, mass is
  the *sum of t-values within it*. Not pixel count. This means a
  small cluster with very strong t-values can have the same mass
  as a large cluster with weak t-values; the metric mixes
  *extent* (how big) and *magnitude* (how strong). Cluster mass
  was introduced by Maris & Oostenveld (2007) and is the dominant
  choice in EEG / MEG / fMRI cluster-permutation work.
- **Stratified label permutation**. To build the null distribution,
  we randomly relabel which producers belong to A vs. B many
  times, recompute the t-map, find clusters, and record the
  *maximum* cluster mass per permutation. "Stratified" means we
  preserve the original group sizes `(N_A, N_B)` exactly each
  time, so the Welch denominator is not perturbed under H0. With
  `paired = TRUE`, the permutation is sign-flip on matched pairs
  instead.
- **Max-statistic FWER**. Family-wise error rate = the probability
  of any false positive across the *entire image*. By comparing
  observed cluster mass to the *maximum* mass under each
  permutation, we directly bound that probability at `alpha`. No
  per-pixel correction is needed; the cluster-level p-value is
  already FWE-corrected.

**When to use it.** "Where in the image do these two conditions
differ, and how confident am I that the differences are not noise?"

```{r rel-cluster, results = "hold"}
clust <- rel_cluster_test(
  sig_A_loc, sig_B_loc,
  img_dims          = c(32L, 32L),
  n_permutations    = 500L,
  cluster_threshold = 2.0,
  alpha             = 0.05,
  seed              = 1L,
  progress          = FALSE
)
print(clust)
```

```{r plot-cluster, fig.cap = "Pixel-wise Welch t-map (red = condition A larger, blue = condition B larger) with significant cluster contours overlaid in black. The two clusters correspond to the two injected signal patches, and only those clusters are FWE-corrected significant at alpha = 0.05; the surrounding noise does not produce false-positive clusters thanks to the max-statistic correction."}
plot(clust)
```

**Reading the result.** `$observed_t` is the raw t-map.
`$clusters` is a data.frame with one row per candidate cluster,
columns: `cluster_id`, `direction` (`"pos"`/`"neg"`), `mass` (sum
of t-values), `size` (pixel count), `p_value` (mass percentile
against the null), `significant` (logical). `$pos_labels` /
`$neg_labels` are integer matrices the same shape as the image,
where each non-zero value tags a cluster (useful for custom
plotting overlays). `$null_distribution` exposes the per-permutation
max masses.

**TFCE alternative.** For threshold-free cluster enhancement,
which integrates across many cluster-forming thresholds and
removes the arbitrary cutoff choice, set `method = "tfce"`. See
§6.7 below for the agreement-map relative.

```{r rel-cluster-tfce, fig.cap = "TFCE-enhanced map (no cluster-forming threshold to choose). Significant pixels are outlined in black. Compared with the threshold-based version above, TFCE generally has more power for diffuse, low-amplitude signal because it integrates evidence across many thresholds rather than committing to one."}
clust_tfce <- rel_cluster_test(
  sig_A_loc, sig_B_loc,
  img_dims        = c(32L, 32L),
  method          = "tfce",
  n_permutations  = 500L,
  tfce_n_steps    = 50L,
  alpha           = 0.05,
  seed            = 1L,
  progress        = FALSE
)
plot(clust_tfce)
```

**Common mistakes.**

- Reading `cluster_threshold = 2.0` as "p < 0.05". It is the
  t-cutoff for forming candidate clusters, not the cluster
  significance threshold.
- Trusting cluster p-values with `n_permutations < 1000` (the
  function warns below 100). Tail probabilities below 0.01 need
  more permutations than that.
- Using `pixel_t_test()` standalone for inference (§6.1).

**Real example (Oliveira et al., 2019, Study 1).** The Dominant vs
Competent contrast on 20-producer signal matrices (256x256 raw masks)
yields spatially-localized, FWE-controlled clusters where the two
group CIs disagree:

```{r cluster-oliveira-2019, eval = FALSE}
rel_cluster_test(sm_dominant, sm_competent,
                 img_dims       = c(256L, 256L),
                 cluster_threshold = 2.0,
                 n_permutations = 1000L,
                 seed           = 1L)
```

```{r oliveira-cluster-fig, echo = FALSE, out.width = "70%", fig.cap = "Cluster test, Dominant vs Competent group CIs (Oliveira et al., 2019, Study 1). Pixel-wise Welch t-map rendered as a translucent diverging overlay on the male base face (positive/red = Dominant larger, negative/blue = Competent larger). Significant clusters at FWE-corrected alpha = .05 are outlined in black."}
fig_DC <- system.file(
  "extdata", "oliveira_2019_examples",
  "cluster_dominant_vs_competent.png",
  package = "rcicrely"
)
if (nzchar(fig_DC)) knitr::include_graphics(fig_DC)
```

Pair this with `rel_dissimilarity()` (next section) for the
overall-magnitude part of the same claim.

### 6.3 `rel_dissimilarity()` {#section-dissim}

**What it does.** Two scalar dissimilarity metrics between the two
conditions' group CIs - Pearson correlation and Euclidean distance -
each with percentile bootstrap CIs.

**When to use it.** "How different are these two conditions overall?"
with uncertainty. Pair with `rel_cluster_test()` for spatial
information.

```{r rel-dissim}
diss <- rel_dissimilarity(
  sig_A, sig_B,
  n_boot   = 500L,
  ci_level = 0.95,
  seed     = 1L,
  progress = FALSE
)
print(diss)
```

```{r plot-dissim, fig.cap = "Bootstrap distributions for Pearson r and Euclidean distance."}
plot(diss)
```

**Reading the result.** `$correlation`, `$euclidean` are the
observed values; `$ci_cor`, `$ci_dist` the percentile CIs;
`$boot_cor`, `$boot_dist` the full bootstrap distributions.

**Common mistakes.** Reading `$correlation` near 1 as "the conditions
look the same" - identical spatial patterns with different magnitudes
give r ~ 1; the magnitude difference shows up in `$euclidean`.
`$euclidean` scales with sqrt(n_pixels) so it is a within-study
metric, not comparable across resolutions.

**Real example (Oliveira et al., 2019, Study 1).** Euclidean
distances between group-mean CIs on 256x256 raw masks, 500
bootstrap replicates each:

```{r dissim-oliveira-2019, eval = FALSE}
rel_dissimilarity(sm_trust,    sm_friendly, n_boot = 500L, seed = 1L)
#> Euclidean = 16.81, 95% CI [19.26, 24.86], normalised = 0.0657
rel_dissimilarity(sm_dominant, sm_competent, n_boot = 500L, seed = 2L)
#> Euclidean = 21.08, 95% CI [21.53, 30.75], normalised = 0.0823
```

The Dominant-vs-Competent contrast yields a substantially larger
Euclidean distance than Trust-vs-Friendly, the same direction as
the published group-CI Pearson correlations (.69 for Trust x
Friendly, -.22 for Dominant x Competent in Table 2 of the paper).
The two metrics report the same pattern from different angles:
Pearson is the within-shape question, Euclidean is the
how-far-apart question.

**Comparing multiple contrasts side-by-side.** When you have
several `rel_dissimilarity()` results and want one figure that
shows whether their CIs overlap, use `compare_dissimilarity()`. It
draws each contrast as a horizontal point with its 95% CI and the
underlying bootstrap density, so the reader can read overlap (or
non-overlap) at a glance. Useful for paper figures.

```{r compare-dissim, eval = FALSE}
d_TF <- rel_dissimilarity(sm_trust,    sm_friendly, n_boot = 500L, seed = 1L)
d_DC <- rel_dissimilarity(sm_dominant, sm_competent, n_boot = 500L, seed = 2L)
compare_dissimilarity(
  "Trust vs Friendly"     = d_TF,
  "Dominant vs Competent" = d_DC
)
```

### 6.4 `run_between()` {#section-run-between}

**What it does.** Calls both between-condition metrics and bundles
the result.

```{r run-between}
between_AB <- run_between(
  sig_A, sig_B,
  img_dims        = c(32L, 32L),
  n_permutations  = 500L,
  n_boot          = 500L,
  seed            = 1L,
  progress        = FALSE
)
print(between_AB)
```

### 6.5 `infoval()` - producer-level signal vs. chance {#section-infoval}

**What it does.** For each producer, compares the Frobenius norm of
their raw mask to a reference distribution simulated from random
sign-weighted aggregations of the same noise pool, at that
producer's trial count. Returns a per-producer z-score against the
reference's median / MAD, plus the full reference distribution and
metadata. A single function handles both 2IFC and Brief-RC; the
difference lives entirely in what you pass as `noise_matrix`.

**When to use it.** As a complement to the `rel_*()` reliability
checks. `infoval()` answers *does each producer's mask carry more
signal than chance?*; `rel_*()` answers *how reliable / how
distinguishable is the group-level pattern?*. They report different
things and a study should usually carry both.

```{r infoval-example, eval = FALSE}
# Brief-RC route: infoval() takes the raw signal_matrix from
# ci_from_responses_briefrc() and the noise_matrix used to generate
# the trials.
res <- ci_from_responses_briefrc(responses, rdata_path = "rcic.Rdata",
                                 base_image_path = "base.jpg")
trial_counts <- with(responses,
                     setNames(table(participant_id),
                              names(table(participant_id))))
iv <- infoval(res$signal_matrix, noise_matrix,
              trial_counts = trial_counts,
              iter = 10000L, cache_path = "infoval_cache.rds",
              seed = 1L)
print(iv)
```

**Reading the result.** `$infoval` is a per-producer z-score vector;
`z > 1.96` is roughly one-tailed *p* < .025 against the chance
reference. `$norms` is the raw observed Frobenius norm per producer.
`$reference` holds the simulated reference norm vectors keyed by
trial count. `$ref_median` and `$ref_mad` are the centre / spread
used for z computation.

#### Troubleshooting low or negative `infoval` z-scores

If you compute `infoval()` and find that *most* or *all* per-producer
z-scores sit well below 1.96 (sometimes negative) even though spot
checks suggest producers are doing the task seriously, that is a
common pattern, not a sign the data is bad. Five reasons in order
of how often they apply:

1. **Frobenius norm is a global energy statistic.** It sums squared
   pixel deviations across the *entire* image. Real internal
   representations are usually spatially sparse (eyes, mouth, jaw,
   maybe 10-30% of pixels), so the 70-90% of "background" pixels
   contribute noise of similar magnitude to the chance reference and
   dilute the signal-bearing region. A producer with strong,
   visually-obvious signal in the eyes can have a Frobenius norm
   only marginally above the random reference.

2. **The reference is strict because it lives in the same subspace.**
   Both the observed mask and the reference are projections onto the
   same low-dimensional sinusoidal noise basis. The reference
   distribution has plenty of overall energy by construction, so the
   only way to clear z = 1.96 is to align signs with a *specific
   subset* of patterns more than chance.

3. **Per-trial signal is small.** Each 2IFC choice contributes a
   tiny signal increment relative to the per-trial noise amplitude.
   With 300 trials the SNR gain is √300 ≈ 17×, but if per-trial
   signal is ~5% of per-trial noise, post-aggregation effective SNR
   is barely visible to a global energy measure.

4. **Without a face mask, infoVal counts background.**
   `face_mask()` ships an oval mask that approximates the Schmitz
   2024 face region. Applying it (`infoval(..., mask =
   face_mask(c(256, 256)))`) concentrates the norm on signal-bearing
   pixels and typically lifts z-scores noticeably.

5. **Group-level CIs have much higher z than individual CIs.**
   Averaging 20 producers' masks reduces noise by √20 ≈ 4.5×, so
   the group-mean CI's effective trial count is `300 × 20 = 6000`
   for a 20-producer condition. The infoVal of the group-mean CI
   (with `trial_counts = setNames(300L * n_producers, "group")`)
   is usually 5-10× the per-producer median; this is a structural
   √N consequence of averaging, not a defect of the per-producer
   metric. Brinkman et al. (2019) themselves only ever computed
   infoVal on *individual* CIs (mean per-producer infoVal in their
   empirical 2IFC data was 3.9 lab / 2.9 online, with 68% / 54%
   of producers individually exceeding 1.96). For group-level
   reporting they recommended inspecting *the distribution of
   per-producer infoVals contributing to the group CI* rather
   than computing one infoVal on the group-mean CI. Either choice
   is defensible; we show the group-mean CI z below as the
   simplest single number, but the distributional view is what
   the 2019 paper endorses.

##### Diagnostic recipe

```{r infoval-diagnostic, eval = FALSE}
sm     <- res$signal_matrix
tc     <- setNames(rep(300L, ncol(sm)), colnames(sm))

# 1. Compare the observed and reference norm distributions
iv  <- infoval(sm, noise_matrix, tc, iter = 1000L,
               cache_path = "iv.rds", seed = 1L)
ref <- iv$reference[[as.character(tc[1])]]
cat(sprintf("observed median = %.4f, reference median = %.4f, %% above = %+.1f%%\n",
            median(iv$norms), median(ref),
            100 * (median(iv$norms) - median(ref)) / median(ref)))

# 2. Apply the face mask
fm  <- face_mask(c(256L, 256L))
iv_masked <- infoval(sm, noise_matrix, tc, mask = fm,
                     iter = 1000L, cache_path = "iv_masked.rds",
                     seed = 1L)
median(iv_masked$infoval)        # typically higher than iv

# 3. Group-mean CI (the headline number to report)
group   <- matrix(rowMeans(sm), ncol = 1, dimnames = list(NULL, "group"))
tc_grp  <- setNames(300L * ncol(sm), "group")
iv_grp  <- infoval(group, noise_matrix, tc_grp, iter = 1000L,
                   seed = 1L)
iv_grp$infoval                   # usually large (e.g. > 5)

# 4. Sanity check with a simulated random responder
set.seed(0)
random_resp <- sample(c(-1, 1), 300L, replace = TRUE)
random_mask <- (noise_matrix %*% random_resp) / 300
iv_rand <- infoval(matrix(random_mask, ncol = 1,
                          dimnames = list(NULL, "rand")),
                   noise_matrix,
                   setNames(300L, "rand"),
                   iter = 1000L, seed = 1L)
iv_rand$infoval                  # should be ~ 0 within MAD noise
```

**Negative z values** mean the observed mask has *less* total
Frobenius energy than the chance reference. This is not "negative
signal"; it is consistent with a producer whose responses align
with a structured low-rank template, since aligning consistently
with one template can reduce the effective dimensionality of the
mask and shrink its overall norm. A negative z is therefore
informative: a clearly negative z (say, < −2) on a producer who
otherwise shows good behaviour usually points to one of:

- The producer responded coherently but to a *different* feature
  than the trait you intended (idiosyncratic representation).
- The reference distribution is mis-calibrated for your design (see
  the v0.2.1 calibration note in the `?infoval` help page;
  early v0.2.0 versions sampled stim ids with replacement and
  inflated reference norms, biasing per-producer z below zero on
  cooperative producers).
- Per-producer trial count is too low to beat chance for any
  meaningful statistic, even with real signal; fall back to the
  group-mean infoVal as the publishable number.

**Reporting recommendation.** Report **(i)** the group-mean
infoVal z and **(ii)** the proportion of producers whose
*individual* z clears 1.96 (with the face mask). Do not report only
the median per-producer z and treat values < 1.96 as evidence of
"no signal"; that conflates the per-producer noisiness of the
metric with the group-level question of whether the CI carries
information.

**Real example (Oliveira et al., 2019, Study 1, Trust condition).**
The same diagnostic recipe above, run on real published data. The
20 individual producers were each shown 300 trials (so the
trial-count-matched reference is simulated at 300):

```{r infoval-oliveira-2019, eval = FALSE}
# Per-producer infoVal
tc <- setNames(rep(300L, 20L), colnames(sm_trust))
iv <- infoval(sm_trust, noise_matrix, tc, iter = 1000L, seed = 1L)
#> per-producer: median z = 0.35, range [-1.69, 2.30], n above 1.96 = 2/20

# Group-mean CI infoVal (effective n_trials = 300 x 20 = 6000)
group  <- matrix(rowMeans(sm_trust), ncol = 1,
                 dimnames = list(NULL, "TRUST_group"))
iv_grp <- infoval(group, noise_matrix,
                  setNames(6000L, "TRUST_group"),
                  iter = 1000L, seed = 1L)
#> group-mean CI z = 11.39

# Face-mask lift on per-producer z
iv_masked <- infoval(sm_trust, noise_matrix, tc,
                     mask = face_mask(c(256L, 256L)),
                     iter = 1000L, seed = 1L)
#> masked median = 0.50  (vs unmasked median = 0.35; lift = +0.15)
```

This is exactly the pattern described in the troubleshooting list
above: per-producer z is centred around 0.35 with only 2 of 20
producers individually above the 1.96 threshold, and yet the
group-mean CI z is 11.39, *unmistakable* signal. The published
Trust/Friendly correlation of .69 (Table 2 of the paper) and the
observable trait dissociations in the group CIs (Figure 1 of the
paper) are both consistent with the group-level z, not with reading
the per-producer median as a "weak signal" verdict. Reporting
recommendation in practice: "the group-mean classification image
showed strong informational value (z = 11.39, n_eff = 6000); 2 of
20 producers individually exceeded z = 1.96."

> **Calibration note.** This output reflects the v0.2.1 fix in
> `simulate_reference_norms()` (sample stim ids without replacement
> when `n_trials <= n_pool`). v0.2.0 over-inflated the reference
> norm and gave `median z = -1.54` for the same data (a bug
> uncovered while replicating Oliveira et al. (2019), Study 1). If
> you computed infoVal under v0.2.0 and got systematically negative
> z-scores, re-run after upgrading.

### 6.6 Region-by-region analyses with `face_mask()` {#section-face-mask}

`face_mask()` accepts a `region` argument that returns a logical
vector for one of: `"full"` (default), `"eyes"`, `"nose"`,
`"mouth"`, `"upper_face"`, `"lower_face"`. The companion
[`load_face_mask()`](#section-load-face-mask) reads an externally-
generated mask (e.g., from `webmorphR::mask_oval()` or a
hand-painted PNG) into the same logical-vector format. Pass either
to `infoval()` directly via the `mask` argument, or apply it via
row-subsetting on a signal matrix to compute any `rel_*()` metric
on a single anatomical region:

> **Apply the mask symmetrically.** This is the single most
> important rule when working with masks in any reliability or
> discriminability analysis: **the same mask must enter both arms
> of any comparison**. For [`infoval()`](#section-infoval), pass
> the mask via the `mask` argument so observed and reference
> Frobenius norms are restricted to the same pixels. For `rel_*()`
> reliability metrics, subset the signal matrix
> (`signal_matrix[mask, , drop = FALSE]`) and compute *all* metrics
> on the same subsetted matrix. For
> [`rel_cluster_test()`](#section-cluster) /
> [`rel_dissimilarity()`](#section-dissim), subset *both*
> condition matrices identically before passing them in. Mixing a
> masked observed value with an unmasked reference (or two
> conditions masked differently) yields a number with no defensible
> interpretation.

```{r face-mask-regions, eval = FALSE}
img <- c(256L, 256L)
m_full  <- face_mask(img, region = "full")
m_eyes  <- face_mask(img, region = "eyes")
m_mouth <- face_mask(img, region = "mouth")

# Region-restricted ICC: pass the masked rows of the signal matrix
rel_icc(sm_trust[m_eyes,  , drop = FALSE])
rel_icc(sm_trust[m_mouth, , drop = FALSE])

# Region-restricted infoVal: pass the mask directly
infoval(sm_trust, noise_matrix,
        trial_counts = setNames(rep(300L, 20L), colnames(sm_trust)),
        iter = 1000L, mask = m_eyes,  seed = 1L)
infoval(sm_trust, noise_matrix,
        trial_counts = setNames(rep(300L, 20L), colnames(sm_trust)),
        iter = 1000L, mask = m_mouth, seed = 1L)
```

Region geometries are heuristic approximations matched to a typical
centred face on a square base image. For non-default base images,
tune the full-face oval via `centre`, `half_width`, `half_height`;
sub-region positions scale relative to that ellipse.

**Real example (Oliveira et al., 2019, Study 1, Trust condition).**
Region-restricted ICC(3,1) on the same 20-producer signal matrix
shows where producer agreement is concentrated:

```{r face-mask-icc-real, eval = FALSE}
# rel_icc(sm_trust[face_mask(c(256L,256L), region = R), , drop = FALSE])
#> region        n_pixels   ICC(3,1)   ICC(3,k)
#> full           32,192     0.109      0.710
#> eyes            1,022     0.069      0.595
#> nose              770     0.028      0.363
#> mouth             966     0.092      0.670
#> upper_face     16,096     0.087      0.655
#> lower_face     16,096     0.130      0.749
```

Producer agreement is strongest in the lower-face region (mouth +
jaw), weakest in the nose. The pattern aligns with the visible
features of the published Trust group CI (Figure 1 of the paper).

Region-restricted infoVal on the same data:

```{r face-mask-infoval-real, eval = FALSE}
# infoval(sm_trust, noise_matrix, tc, mask = face_mask(c(256L,256L), region = R), ...)
#> region   median z   range          n above 1.96
#> full     +0.51      [-1.81, +3.13]      4/20
#> eyes     +0.55      [-0.98, +1.56]      0/20
#> mouth    +0.52      [-0.76, +4.67]      3/20
```

Regional infoVal works as a sensitivity check: if a region's
median z is markedly lower than the full-face median, the group
signal is *not* concentrated in that region; if it is markedly
higher, the region carries more than its pixel-share of the signal.
For Trust, the eye and mouth regions both replicate the full-face
median; the eye region's per-producer range is narrower, suggesting
producers agree on *direction* in the eye region but the energy is
modest.

Region-restricted **between-condition** inference works the same
way: zero-fill outside the region in both signal matrices and pass
them to `rel_cluster_test()`, so clusters can only form inside the
kept region. Here is Trustworthy vs Dominant, restricted to the eyes
and mouth:

```{r cluster-region-trust-dominant, eval = FALSE}
img_dims <- c(256L, 256L)
m_eyes   <- face_mask(img_dims, region = "eyes")
m_mouth  <- face_mask(img_dims, region = "mouth")

zero_fill <- function(sm, keep) {
  out <- sm
  out[!keep, ] <- 0
  out
}

clust_eyes  <- rel_cluster_test(
  zero_fill(sm_trust,    m_eyes),
  zero_fill(sm_dominant, m_eyes),
  img_dims          = img_dims,
  cluster_threshold = 2.0,
  n_permutations    = 1000L,
  seed              = 1L
)

clust_mouth <- rel_cluster_test(
  zero_fill(sm_trust,    m_mouth),
  zero_fill(sm_dominant, m_mouth),
  img_dims          = img_dims,
  cluster_threshold = 2.0,
  n_permutations    = 1000L,
  seed              = 1L
)
```

```{r oliveira-region-fig, echo = FALSE, out.width = "95%", fig.cap = "Region-restricted cluster test, Trustworthy vs Dominant group CIs (Oliveira et al., 2019, Study 1). Pixel-wise Welch t-map rendered as a translucent diverging overlay on the male base face; positive/red = Trustworthy larger, negative/blue = Dominant larger. Eyes region (left panel) and mouth region (right panel). Significant clusters at FWE-corrected alpha = .05 are outlined in black."}
fig_TD <- system.file(
  "extdata", "oliveira_2019_examples",
  "cluster_trust_vs_dominant_eyes_mouth.png",
  package = "rcicrely"
)
if (nzchar(fig_TD)) knitr::include_graphics(fig_TD)
```

**Critical-review note.** Region masks are heuristic. They do not
account for individual differences in face geometry (the base
face's eyes, nose, mouth occupy different pixels on different
photographs), and they assume the typical centred-face layout
shipped with most rcicr stimulus pools. For studies using
non-centred or differently-proportioned base faces, define a custom
mask via `ellipse_mask()` (internal) or by hand-curating a logical
pixel vector. The geometries shipped here are calibrated to KDEF-
style 256x256 male faces and should reproduce reasonably for
most published RC datasets.

### 6.7 Visualising producer agreement: `plot_agreement_map()`

A complement to the group-mean CI image: a heatmap of where
producers in a single condition *agree* on the direction of signal.
Each pixel's colour is proportional to a one-sample t-statistic
across producers (mean / standard error), so positive saturation
means agreement on positive signal and negative saturation means
agreement on negative signal.

```{r plot-agreement-map, eval = FALSE}
# Continuous t-map (default)
plot_agreement_map(sm_trust)

# Region-restricted, mouth only
plot_agreement_map(sm_trust,
                   mask = face_mask(c(256L, 256L), region = "mouth"))

# Thresholded view (only pixels with |t| > 2.0 are coloured)
plot_agreement_map(sm_trust, threshold = 2.0)
```

Use this for **descriptive** figures of where producers agree
within a condition; pair with `rel_cluster_test()` (which is the
**inferential** counterpart for between-condition pixel-level
contrasts).

## 7. Brief-RC end-to-end

Schmitz, Rougier & Yzerbyt (2024) introduced Brief-RC 12: each trial
shows 12 noisy faces (6 oriented, 6 inverted, drawn from a shared
noise pool); participant picks one. The recorded format is one row
per trial, columns `participant_id`, `trial`, `stimulus` (chosen pool
id), `response` (`+1` if oriented, `-1` if inverted). `rcicrely`'s
implementation is native (no rcicr `_brief` functions, which don't
exist in the upstream rcicr v1.0.1).

```{r briefrc-demo, eval = FALSE}
# 1. Generate the noise pool (once per project) - heavy step.
rcicr::generateStimuli2IFC(
  base_face_files     = list(base = "data/base.jpg"),
  n_trials            = 600L,        # becomes the Brief-RC pool size
  img_size            = 256L,
  stimulus_path       = "data/stim",
  label               = "mystudy",
  ncores              = 1L,          # macOS-safe default
  return_as_dataframe = TRUE,        # capture the noise matrix!
  seed                = 1L
) |> saveRDS("data/noise_matrix.rds")  # cache for fast loading

# 2. Collect Brief-RC responses in the per-trial format.

# 3. Compute individual CIs.
res <- ci_from_responses_briefrc(
  responses       = my_brief_rc_responses,
  noise_matrix    = readRDS("data/noise_matrix.rds"),
  base_image_path = "data/base.jpg",
  method          = "briefrc12",
  scaling         = "matched"   # also rendered for plotting
)

# 4. Reliability assessment.
within  <- run_within(res$signal_matrix, seed = 1L)
between <- run_between(res_A$signal_matrix, res_B$signal_matrix,
                       seed = 1L)

# 5. Save rendered CI to PNG (visualization only - DO NOT use for stats)
if (requireNamespace("png", quietly = TRUE)) {
  for (j in seq_along(res$participants)) {
    img <- matrix(res$rendered_ci[, j], res$img_dims[1], res$img_dims[2])
    img <- pmin(pmax(img, 0), 1)
    png::writePNG(img, sprintf("plots/p%02d.png", j))
  }
}
```

**Brief-RC `infoVal`** ships in v0.2 as `infoval()` (see §6.5). The
reference distribution is matched to each producer's recorded trial
count (number of `(stim, +/-1)` contributions entering the mask), not
to the pool size. For 2IFC the two are equal so the original rcicr
default is fine; for Brief-RC each producer typically records fewer
choices than there are pool items, so a pool-size reference biases
infoVal downward. `infoval()` closes that calibration gap. Pass the
**raw mask** (`res$signal_matrix`), not `res$rendered_ci` or a PNG.

Two notes on interpretation. First, "subset" here means the number
of **recorded** choices that enter the mask (one `(stim, +/-1)` per
trial), not the producer's perceptual exposure: Brief-RC trials are
*cognitively* richer than 2IFC trials (12 noisy faces of context vs.
2), so producers do not see less of the pool. The trial-count match
is the right calibration for infoVal's magnitude statistic, not a
claim about cognitive exposure. Second, infoVal measures the
**magnitude** of the mask (Frobenius norm) against a chance reference,
so it does not distinguish a more accurate mask from a larger one.
Treat absolute Brief-RC vs 2IFC z-score differences with caution;
`rel_split_half()`/`rel_icc()` (signal stability) and
`rel_cluster_test()`/`rel_dissimilarity()` (between-condition
separability) are the right complements when pattern accuracy matters
more than magnitude.

## 8. Interpreting results and common pitfalls

**Sample size.** `rcicrely` aborts below 4 producers per condition
and warns below 30; a sample of N >= 60 producers is recommended for
stable reliability assessment. At N = 10 the metrics are computable
but their CIs are wide.

**Split-half.** Think of `r_hh` ~ 0.3 as weak, 0.5 as moderate, 0.7+
as strong producer-level consistency. `r_SB` is what to report when
asked about the full-sample CI.

**LOO flagging.** A flag is a starting point for investigation, not a
verdict. Use `flag_method = "mad"` if the SD rule is dominated by
the very outliers it tries to find.

**ICC(3,*) vs ICC(2,*).** Numerically close at high pixel counts;
ICC(3,*) is the correctly-specified model (pixels fixed, participants
random). Use `variants = c("3_1", "3_k", "2_1", "2_k")` when both
are needed.

**Cluster threshold != p-value.** `cluster_threshold = 2.0` is the
t-cutoff for forming clusters. Significance is decided by mass
relative to the null. The default is roughly two-tailed p < 0.05 at
large df, but at small N it under-rejects - inspect the t histogram.

**Raw vs. rendered CIs.** This package's reliability math is correct
**given** a raw signal matrix. Mode 1 (PNGs on disk) loads
`scaling(mask)` instead of `mask`; results will differ from what raw
responses would produce. The size of the difference depends on the
scaling that was applied at PNG-write time. Mode 2 is the safe path.

**`infoVal` correctness.** `rcicr::computeInfoVal2IFC()` extracts
the raw `$ci` from its input CI-list internally, so the standard
2IFC path is safe regardless of `scaling`. The pitfall is in
hand-rolled `infoVal` implementations, for Brief-RC, or for any
custom code that consumes a bare matrix, which must be fed the
raw mask. Feeding rendered/scaled values silently distorts the
reference comparison and produces wrong significance verdicts.

## 9. References

- Brinkman, L., Todorov, A., & Dotsch, R. (2017). Visualising mental
  representations: A primer on noise-based reverse correlation in
  social psychology. *European Review of Social Psychology*, 28(1),
  333-361. <https://doi.org/10.1080/10463283.2017.1381469>
- Brinkman, L., Goffin, S., van de Schoot, R., van Haren, N. E. M.,
  Dotsch, R., & Aarts, H. (2019). Quantifying the informational
  value of classification images. *Behavior Research Methods*,
  51(5), 2059-2073. <https://doi.org/10.3758/s13428-019-01232-2>
- Cone, J., Brown-Iannuzzi, J. L., Lei, R., & Dotsch, R. (2021). Type I Error Is Inflated in the Two-Phase Reverse Correlation Procedure. *Social Psychological and Personality Science*, 12(5), 760-768. <https://doi.org/10.1177/1948550620938616>
- Dotsch, R. (2023). *rcicr*: Reverse-Correlation Image-Classification
  Toolbox. R package v1.0.1.
  <https://github.com/rdotsch/rcicr>
- Maris, E., & Oostenveld, R. (2007). Nonparametric statistical
  testing of EEG- and MEG-data. *Journal of Neuroscience Methods*,
  164(1), 177-190. <https://doi.org/10.1016/j.jneumeth.2007.03.024>
- McGraw, K. O., & Wong, S. P. (1996). Forming inferences about some
  intraclass correlation coefficients. *Psychological Methods*,
  1(1), 30-46. <https://doi.org/10.1037/1082-989X.1.1.30>
- Nichols, T. E., & Holmes, A. P. (2002). Nonparametric permutation
  tests for functional neuroimaging: A primer with examples.
  *Human Brain Mapping*, 15(1), 1-25.
  <https://doi.org/10.1002/hbm.1058>
- Oliveira, M., Garcia-Marques, T., Dotsch, R., & Garcia-Marques, L.
  (2019). Dominance and competence face to face: Dissociations
  obtained with a reverse correlation approach. *European Journal
  of Social Psychology*. <https://doi.org/10.1002/ejsp.2569>.
  Open data and analysis scripts: <https://doi.org/10.17605/osf.io/hr5pd>.
- Schmitz, M., Rougier, M., & Yzerbyt, V. (2024). Introducing the
  brief reverse correlation: An improved tool to assess visual
  representations. *European Journal of Social Psychology*.
  <https://doi.org/10.1002/ejsp.3100>
- Shrout, P. E., & Fleiss, J. L. (1979). Intraclass correlations:
  Uses in assessing rater reliability. *Psychological Bulletin*,
  86(2), 420-428. <https://doi.org/10.1037/0033-2909.86.2.420>

[rcicrdiagnostics-link]: https://github.com/olivethree/rcicrdiagnostics
